---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vegasr

<!-- badges: start -->
<!-- badges: end -->

Vegasr is an R library for **multi-dimensional numerical integration**. It is a thin wrapper around the Python [vegas](https://vegas.readthedocs.io/en/latest/index.html) library which was developed and maintained by the author of the vegas algorithm, G. P. Lepage. The purpose of this library is to allow ready access to the latest [vegas+](https://arxiv.org/abs/2009.05112) algorithm from R.  

The vegas algorithm performs efficient Monte Carlo integration for integrals of modest to high dimension where cubature or hcubature methods are practically infeasible. The [vegas+](https://arxiv.org/abs/2009.05112) algorithm is a more efficient recent variant of the same author's original and widely used vegas algorithm. For the original vegas article see J. Comput. Phys. 27 (1978) 192 and for later vegas+ see J. Comput. Phys. 439 (2021) 110386)

## Installation

An essential pre-requisite is Python must be available. The repo is [here](https://github.com/fraseriainlewis/vegasr)

You can install the development version of vegasr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("fraseriainlewis/vegasr")
vegas_install() # this is needed in addition to install python environment
```

## Example 1

This is a basic example to show the integrand function structure and core functionality. 

```{r example1a}
library(vegasr)
## Important - run next line in each R session to ensure python is ready
vegas_initialize()

# Use simple example of 3-D Multivariate Normal Density
library(mvtnorm) 
mu<-matrix(c(0.5, -0.2, 0.1),nrow=1) 
cov<-matrix(data=c(
                 1.0, 0.5, 0.2,
                 0.5, 1.2, 0.3,
                 0.2, 0.3, 0.8),ncol=3,byrow=FALSE)

# the integrand function MUST take a matrix of dimension [BATCH,M] as first 
# argument and return a vector of length M. Another other additional 
# named arguments allowed. See ?vegas_integrate for more details.

myf<-function(x,mu,cov){
  res<-dmvnorm(x,mean = mu,sigma=cov)
  return(res)
  }
```

Now run the vegas+ algorithm to achieve a tolerance of 0.1% error. 

```{r example1b}
## See help page for descriptions of warm and nitn and neval.
result<-vegas_integrate(f=myf,
                        lower=c(-5,-5,-5), upper=c(5,5,5),
                        nitn_warm = 10, neval_warm = 10000,
                        nitn = 10, neval = 10000,
                        errTol=0.1,maxIter=20,
                        mu=mu,cov=cov) # these are additional arguments needed for myf
print(result)
#cat("Estimate = ",result$mean," error = ",result$error,"\n")

```

# Example 2
This example is 5 dimensions and has a known numerical solution as a validation check. 
```{r example2}
library(vegasr)
## Important - run next line in each R session to ensure python is ready
vegas_initialize()

# 5-D Gaussian density

# true numerical value using pnorm()
alpha<-matrix(data=c(1.),nrow=1)
erf<-function(x){return(2 * pnorm(x * sqrt(2)) - 1)}
oneD<-sqrt(pi / alpha[1]) * erf(sqrt(alpha[1]) / 2.0)
true.val<-oneD**5
print(true.val)

# now define custom integrand
gaus<-function(x,alpha){
  # x shape is (batch, dim)
  # Compute sum of squared differences from 0.5
  dx2 = apply((x-0.5)**2,1,sum)
  return(exp(-alpha[1] * dx2))
}

# check that gaus(x,alpha) does work with the correct shapes
myx<-matrix(data=rep(0.7,10),ncol=5) # 2 row 5 col
gaus(x=myx,alpha=alpha) # returns length 2 vector

# now do the Vegas estimation
result2<-vegas_integrate(f=gaus,
                        lower=rep(0,5), upper=rep(1,5),
                        nitn_warm = 10, neval_warm = 10000,
                        nitn = 5, neval = 10000,
                        errTol=0.01,maxIter=100,
                        alpha=alpha)
print(result2)

cat("using pnorm() = ", true.val, " using vegas = ", result2$mean,"\n")
```
