---
title: "BHM Basket Trial Logistic Regression: RStan vs vegasr"
from: markdown+tex_math_dollars
format:
  html:
    toc: false
    embed-resources: true
    html-math-method: mathjax
---

```{r Setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```


## Quick start
This vignette is almost identical to ``BHM Basket Trial Logistic Regression Vegas v RStan'' but uses Apple's MLX library so the computations are offloaded onto a GPU. This results in a considerable speed up in terms of computational time.  

### Model Formulation

The model implemented here is a non-centered parameterization for a Basket trial with two baskets and a logistic response model.

$$
\begin{aligned}
\mu_0 &\sim \text{Normal}(0, 2.5)\\
\sigma_0 &\sim \text{Half-Normal}(0, 2.5) \\
\mu_1 &\sim \text{Normal}(0, 2.5) \\
\sigma_1 &\sim \text{Half-Normal}(0, 2.5) \\
\tilde{\theta}_j &\sim \text{Normal}(0, 1)\quad\enspace\text{for}\enspace{j=1,\dots,4}  \\
\beta_{0,0} &= \mu_0 + \sigma_0 \tilde{\theta}_1 \quad\quad\hspace{0.2cm}\text{ intercept - log odds control basket 0}\\
\beta_{0,1} &= \mu_0 + \sigma_0 \tilde{\theta}_2 \quad\quad\hspace{0.2cm}\text{ intercept - log odds control basket 1}\\
\beta_{1,0} &= \mu_1 + \sigma_1 \tilde{\theta}_3 \quad\quad\hspace{0.2cm}\text{ treatment - log OR basket 0}\\
\beta_{1,1} &= \mu_1 + \sigma_1 \tilde{\theta}_4 \quad\quad\hspace{0.2cm}\text{ treatment - log OR basket 1}\\
\text{logit(}{p_i}) &= \beta_0 + \beta_1 z_i\quad\quad\enspace\enspace\text{for}\enspace{i=1,\dots,N}  \\
y_i &\sim \text{Bernoulli}(p_i)
\end{aligned}
$$

When using direct integration via Vegas we don't need the non-centered parameterization:


$$
\begin{aligned}
\beta_{0,0} &\sim \text{Normal}(\mu_0, \sigma_0) \\
\beta_{0,1} &\sim \text{Normal}(\mu_0, \sigma_0) \\
\beta_{1,0} &\sim \text{Normal}(\mu_1, \sigma_1) \\
\beta_{1,1} &\sim \text{Normal}(\mu_1, \sigma_1)
\end{aligned}
$$

```{r rsetup, include=FALSE,eval=TRUE}
library(rstan)
library(bayesplot)
library(ggplot2)
library(vegasr)
#reticulate::py_install("bash") # just a test
```

## Data set

The R chunk below creates a simple dataset of three cols:

-   a binary response variable (0/1 = non-responder/responder),
-   a binary treatment variable (0/1 = control/test treatment)
-   and a basket ID variable. (1)

The basket ID is currently set fixed at 1, denoting there is only one basket in this trial, i.e. a classical two arm randomized trial design. Baskets will be added in other vignettes.

```{r datasets, include=TRUE,eval=TRUE}
### Prepare model inputs
fn_create_data_2<-function(seed){
# Set up data
rr_k_ctrl <- c(0.20,0.35, 0.25, 0.30, 0.35)        # control response rate for each basket
rr_k_trt <- c(0.40,0.45, 0.45, 0.50, 0.55)         # treatment response rate for each basket

K<-length(rr_k_ctrl)        # number of baskets

N_k_ctrl <- rep(50, K)     # number of control participants per basket
N_k_trt <- rep(50, K)      # number of treatment participants per basket
N_k <- N_k_ctrl + N_k_trt   # number of participants per basket (both arms combined)
N <- sum(N_k)               # total sample size
k_vec <- rep(1:K, N_k)      # N x 1 vector of basket indicators (1 to K)

z_vec<-NULL;
y<-NULL;
for(i in 1:K){ # for each basket repeat 0-control 1-trt according to the specifc Ns
  z_vec<-c(z_vec,rep(0:1,c(N_k_ctrl[i],N_k_trt[i]))) # treatment/control indicator
  y<-c(y,
       c(rbinom(N_k_ctrl[i],1,rr_k_ctrl[i]), # bernoulli for control
         rbinom(N_k_trt[i],1,rr_k_trt[i]))) #           for trt
}

thedata<-data.frame(y,basketID=k_vec,Treatment=z_vec)

# response
  y<-matrix(data=as.numeric(thedata$y),ncol=1)
  # treatment
  treat<-matrix(data=as.numeric(thedata$Treatment),ncol=1)
  # basket
  basketID<-matrix(data=as.numeric(thedata$basketID),ncol=1)
  
  return(list(y=y,treat=treat, basketID = basketID))
}

thedata <- fn_create_data_2(99999)
```


## RStan

The RStan workflow is somewhat more compact than the TFP equivalent, where TFP requires more lower level details to be specified whereas these are dealt with behind the scences in RStan. The nodel definition This is somewhat simplified for ease of use, e.g. hard coded hyperparameters.

```{r stanmodeldef,include=TRUE,echo=TRUE,eval=TRUE}
### Stan model
  BHM_stan_1 <- "data {
  int<lower = 0> K;                     // number of baskets (K >= 2)
  int<lower = 0> N;                     // total number of participants
  //int<lower = 0, upper = N> N_k[K];     // K x 1 vector of basket sample sizes
  //int<lower = 0, upper = N> N_k;
  int<lower = 1, upper = K> k_vec[N];   // N x 1 vector of basket indicators
  int<lower = 0, upper = 1> z_vec[N];   // N x 1 vector of treatment indicators for active treatment arm
  int<lower = 0, upper = 1> y[N];       // N x 1 vector of binary responses
}
parameters {
  matrix[K,2] beta_tr;                  // K x 2 matrix of transformed regression coefficients
  real mu1;                              // scalar of hierarchical mean (log odds ratio)
  real<lower = 0> sigma1;                  // scalar of hierarchical SD (log odds ratio)
  real mu0;                              // scalar of hierarchical mean (log odds ratio)
  real<lower = 0> sigma0;                  // scalar of hierarchical SD (log odds ratio)

}
transformed parameters {
  matrix[K,2] beta;                     // K x 2 matrix of regression coefficients (without transformation)
  // Calculate beta coefs using transformed betas (leads to better mixing)
  for (k in 1:K){
    beta[k,1] = mu0 + sigma0 * beta_tr[k,1];
    beta[k,2] = mu1 + sigma1 * beta_tr[k,2];
  }

}
model {
  mu0 ~ normal(0, 2.5);    // vectorized normal priors for hierarchical means (specify SD in normal distn)
  sigma0 ~ normal(0, 2.5);       // vectorized half-normal priors for hierarchical SDs (specify SD in normal distn)
  mu1 ~ normal(0, 2.5);    // vectorized normal priors for hierarchical means (specify SD in normal distn)
  sigma1 ~ normal(0, 2.5);       // vectorized half-normal priors for hierarchical SDs (specify SD in normal distn)
  beta_tr[,1] ~ normal(0, 1);  // vectorized normal prior for the log odds for the control arm
  beta_tr[,2] ~ normal(0, 1);  // vectorized normal prior for the log odds ratios
  for (i in 1:N)
    y[i] ~ bernoulli_logit(beta[k_vec[i], 1] + beta[k_vec[i], 2] * z_vec[i]);
}
"

```

### Setup and run sampler

```{r stansample,echo=TRUE,results='hide',eval=TRUE}

# Create list with input values for Stan model
data_input_1 <- list(
  K = length(unique(thedata$basketID)),                # number of subgroups
  N = length(thedata$basketID),                # total sample size
  #N_k = N-k,            # sample size per basket
  k_vec = as.integer(thedata$basketID),        # N x 1 vector of subgroup indicators
  z_vec =as.integer(thedata$treat),        # N x 1 vector of active treatment arm indicators
  y = as.integer(thedata$y)                 # N x 1 vector of binary responses
  )



### Compile and fit model
options(mc.cores = parallel::detectCores())
# Compile model (only run the line below once for a simulation study - compilation not dependent on any
# simulation inputs as defined by scenarios or on simulated data)
stan_mod_1 <- stan_model(model_code = BHM_stan_1)
# Fit model
nsamps <- 30000       # number of posterior samples (after removing burn-in) per chain
nburnin <- 1000       # number of burn-in samples to remove at beginning of each chain
nchains <- 4          # number of chains
#BHM_pars_1 <- c("beta","mu0", "sigma0","mu1", "sigma1", "beta","beta_tr")     # parameters to sample
start_time <- Sys.time()
stan_fit_2 <- sampling(stan_mod_1, data = data_input_1,
                       iter = nsamps + nburnin, warmup = nburnin, chains = nchains)
end_time <- Sys.time()
end_time - start_time
post_draws_2 <- as.matrix(stan_fit_2)           # posterior samples of each parameter

```

### Summarize results

```{r stanresults,fig.width=10, fig.height=6, out.width="100%",eval=TRUE}
library(bayesplot)
color_scheme_set("viridisA")
p<-mcmc_trace(stan_fit_2,"mu0")
#ggsave("precomputed/vg2_traceplot1_stan.png", p)
print(p)
```
```{r stanresults2,fig.width=10, fig.height=6, out.width="100%",eval=TRUE}
library(bayesplot)
color_scheme_set("viridisA")
p<-mcmc_trace(stan_fit_2,"beta[1,1]")
#ggsave("precomputed/vg2_traceplot1_stan.png", p)
print(p)
```

## Now for the vegasr Equivalent Model

### Define the log posterior function

```{r logpos1, echo=TRUE, eval=TRUE}
fn_log_post_K <- function(theta, y, treat, basketID, shiftby, uselog, K= NULL) {

  y <- as.numeric(y)
  treat <- as.numeric(treat)
  basketID <- as.integer(basketID)
  

  if (is.null(K)) K <- max(basketID, na.rm = TRUE)
  if (any(basketID < 1L | basketID > K)) stop("basketID must be in 1..K")

  B <- nrow(theta)
  M_expected <- 2*K + 4
  if (ncol(theta) != M_expected) {
    stop(sprintf("theta must have %d columns (2*K+4). Got %d.", M_expected, ncol(theta)))
  }

  # split theta
  theta0 <- pmin(pmax(theta[, 1:K, drop=FALSE],           -0.9999), 0.9999)
  theta1 <- pmin(pmax(theta[, (K+1):(2*K), drop=FALSE],   -0.9999), 0.9999)
  theta2 <- pmin(pmax(theta[, 2*K + 1],                  -0.9999), 0.9999)
  theta3 <- pmin(pmax(theta[, 2*K + 2],                  -0.9999), 0.9999)
  theta4 <- pmin(pmax(theta[, 2*K + 3],                   1e-5),   0.9999)
  theta5 <- pmin(pmax(theta[, 2*K + 4],                   1e-5),   0.9999)

  log_jac_term <- function(th) log1p(th^2) - 2*log1p(-(th^2))

  jacobianLB <- rowSums(log_jac_term(theta0)) +
    rowSums(log_jac_term(theta1)) +
    log_jac_term(theta2) + log_jac_term(theta3) + log_jac_term(theta4) + log_jac_term(theta5)

  a0 <- theta0/(1 - theta0^2)   # B x K
  a1 <- theta1/(1 - theta1^2)   # B x K
  mu0 <- theta2/(1 - theta2^2)  # length B
  mu1 <- theta3/(1 - theta3^2)
  sigma0 <- theta4/(1 - theta4^2)
  sigma1 <- theta5/(1 - theta5^2)

  treat_BN <- matrix(rep(treat, each = B), nrow = B)  # B x N
  eta_BN <- a0[, basketID, drop=FALSE] + a1[, basketID, drop=FALSE] * treat_BN
  eta <- t(eta_BN)  # N x B

  log1pexp <- function(x) ifelse(x > 0, x + log1p(exp(-x)), log1p(exp(x)))
  logL <- colSums(y * eta - log1p(exp(eta)))  # length B

  # priors: dnorm will recycle mu0/sigma0 across columns (K)
  prior_a0 <- rowSums(stats::dnorm(a0, mean = mu0, sd = sigma0, log = TRUE))
  prior_a1 <- rowSums(stats::dnorm(a1, mean = mu1, sd = sigma1, log = TRUE))
  prior_mu0 <- stats::dnorm(mu0, mean = 0, sd = 2.5, log = TRUE)
  prior_mu1 <- stats::dnorm(mu1, mean = 0, sd = 2.5, log = TRUE)
  prior_sigma0 <- extraDistr::dhnorm(sigma0, sigma = 2.5, log = TRUE)
  prior_sigma1 <- extraDistr::dhnorm(sigma1, sigma = 2.5, log = TRUE)

  logPost <- logL + prior_a0 + prior_a1 + prior_mu0 + prior_mu1 + prior_sigma0 + prior_sigma1 + jacobianLB

  if (uselog == 1) return(logPost - shiftby[1])
  exp(logPost - shiftby[1])
}

K <- length(unique(thedata$basketID))

lower <- c(rep(-1, 2*K), -1, -1, 1e-4, 1e-4)
upper <- c(rep( 1, 2*K),  1,  1,   1,   1)

result_logEv <- vegasBayesEvidence(
  f = fn_log_post_K,
  lower = lower, upper = upper,
  nitn_warm = 10, neval_warm = 10000,
  nitn = 10, neval = 10000,
  errTol = 0.1, maxIter = 10, seed = 99999, nsearch = 10000,
  extra_args=list(
    y=thedata$y,
    treat=thedata$treat,
    basketID = thedata$basketID,
    shiftby=0,uselog=1.)
)
cat("log evidence = ",result_logEv,"\n")


```


### Marginalize for a1[,2]

```{r marg_a1_2}

fn_marg_a1_2 <- function(theta, y, treat, basketID, shiftby, uselog, z, K = NULL) {

  y <- as.numeric(y)
  treat <- as.numeric(treat)
  basketID <- as.integer(basketID)
  

  if (is.null(K)) K <- max(basketID, na.rm = TRUE)
  if (K < 2) stop("Need K >= 2 to marginalize a1[,2].")
  if (any(basketID < 1L | basketID > K)) stop("basketID must be in 1..K")

  B <- nrow(theta)

  # We removed theta1[,2], so dimension is (2*K + 3):
  # theta0: K cols
  # theta1_free: (K-1) cols (all baskets except 2)
  # theta2..theta5: 4 scalars
  M_expected <- 2*K + 3
  if (ncol(theta) != M_expected) {
    stop(sprintf("theta must have %d columns (2*K+3). Got %d.", M_expected, ncol(theta)))
  }

  # split theta
  theta0 <- pmin(pmax(theta[, 1:K, drop=FALSE], -0.9999), 0.9999)

  # theta1 for baskets except 2, in order: 1,3,4,...,K
  k_free <- setdiff(seq_len(K), 2L)
  theta1_free <- pmin(pmax(theta[, (K+1):(K+(K-1)), drop=FALSE], -0.9999), 0.9999)

  # remaining scalars (shifted by -1 compared to full model)
  theta2 <- pmin(pmax(theta[, 2*K],     -0.9999), 0.9999)  # mu0
  theta3 <- pmin(pmax(theta[, 2*K + 1], -0.9999), 0.9999)  # mu1
  theta4 <- pmin(pmax(theta[, 2*K + 2],  1e-5),   0.9999)  # sigma0
  theta5 <- pmin(pmax(theta[, 2*K + 3],  1e-5),   0.9999)  # sigma1

  log_jac_term <- function(th) log1p(th^2) - 2*log1p(-(th^2))

  # Jacobian: include theta0, theta1_free, theta2..theta5; exclude removed theta1[,2]
  jacobian <- rowSums(log_jac_term(theta0)) +
    rowSums(log_jac_term(theta1_free)) +
    log_jac_term(theta2) + log_jac_term(theta3) + log_jac_term(theta4) + log_jac_term(theta5)

  a0 <- theta0/(1 - theta0^2)  # B x K

  a1 <- matrix(NA_real_, nrow = B, ncol = K)
  a1[, 2] <- z
  a1[, k_free] <- theta1_free/(1 - theta1_free^2)  # fill other baskets

  mu0 <- theta2/(1 - theta2^2)
  mu1 <- theta3/(1 - theta3^2)
  sigma0 <- theta4/(1 - theta4^2)
  sigma1 <- theta5/(1 - theta5^2)

  treat_BN <- matrix(rep(treat, each = B), nrow = B)  # B x N
  eta_BN <- a0[, basketID, drop=FALSE] + a1[, basketID, drop=FALSE] * treat_BN
  eta <- t(eta_BN)  # N x B

  logL <- colSums(y * eta - log1p(exp(eta)))

  prior_a0 <- rowSums(stats::dnorm(a0, mean = mu0, sd = sigma0, log = TRUE))
  prior_a1 <- rowSums(stats::dnorm(a1, mean = mu1, sd = sigma1, log = TRUE))
  prior_mu0 <- stats::dnorm(mu0, mean = 0, sd = 2.5, log = TRUE)
  prior_mu1 <- stats::dnorm(mu1, mean = 0, sd = 2.5, log = TRUE)
  prior_sigma0 <- extraDistr::dhnorm(sigma0, sigma = 2.5, log = TRUE)
  prior_sigma1 <- extraDistr::dhnorm(sigma1, sigma = 2.5, log = TRUE)

  logPost <- logL + prior_a0 + prior_a1 + prior_mu0 + prior_mu1 + prior_sigma0 + prior_sigma1 + jacobian

  if (uselog == 1) return(logPost - shiftby[1])
  exp(logPost - shiftby[1])
}

basketID_int <- as.integer(factor(thedata$basketID))  # ensure 1..K
K <- max(basketID_int)

lower_full <- c(rep(-1, 2*K), -1, -1, 1e-4, 1e-4)
upper_full <- c(rep( 1, 2*K),  1,  1,   1,   1)

idx_drop <- K + 2  # position of theta1[,2] in the full parameter vector
lower_red <- lower_full[-idx_drop]
upper_red <- upper_full[-idx_drop]

dens_at_z <- vegasBayesPosterior(
  f = fn_marg_a1_2,
  lower = lower_red, upper = upper_red,
  log_evidence = result_logEv,
  extra_args=list(
    y=thedata$y,
    treat=thedata$treat,
    basketID = thedata$basketID,
    shiftby=0,uselog=1.,
    z = 0.91)
)

cat("Marginal density f(z) at z = 0.91", dens_at_z, "\n")

```
```{r}
library(foreach)
library(doParallel)
library(extraDistr)
library(tictoc)

cl <- makeCluster(4)

registerDoParallel(parallel::detectCores() - 2)

# Setting integrand limits based on Stan posterior draws 
a_1_2_draws <- as.vector(rstan::extract(stan_fit_2, "beta[2,2]")$"beta[2,2]")
myz <- seq(quantile(a_1_2_draws, 0.001), quantile(a_1_2_draws, 0.999), length.out = 50)


tic("Parallel Vegas Loop") # Start timer with a label
f_z<-foreach(z= myz,.packages = c("extraDistr", "vegasr")) %dopar% {
  vegasBayesPosterior(f= fn_marg_a1_2,
                      lower = lower_red, upper = upper_red,
                      log_evidence = result_logEv,
                      nitn_warm = 10, neval_warm = 10000,
                      nitn = 10, neval = 100000,
                      errTol=0.1,maxIter=10,seed=99999,nsearch=10000,
                      
                      extra_args = list(
                        y = as.numeric(thedata$y),
                        treat = as.numeric(thedata$treat),
                        basketID = as.integer(thedata$basketID),
                        shiftby = 0, uselog = 1., z = z
                      )
  )
  
}

stopCluster(cl)
toc() # Stops timer and prints
```
### Plot the vegasr beta[2,2] posterior density

```{r plotvegasr, include=TRUE, eval=TRUE, fig.width=10, fig.height=6, out.width="100%", fig.align='center'}

library(ggplot2)

# 1. Prepare data for plotting
# Ensure f_z is a numeric vector (in case it is a list from foreach)
if (is.list(f_z)) {
  # If it's a list of results, extract the 'mean' component
  if (is.list(f_z[[1]]) && "mean" %in% names(f_z[[1]])) {
    f_z_vec <- unlist(lapply(f_z, function(x) x$mean))
  } else {
    f_z_vec <- unlist(f_z)
  }
} else {
  f_z_vec <- f_z
}

# Create a smooth interpolation for the line
z_grid <- seq(min(myz), max(myz), length.out = 500)
f_interp <- splinefun(myz, f_z_vec, method = "fmm")

# Prepare data frames
plot_data <- data.frame(z = as.numeric(myz), f = as.numeric(f_z_vec))
line_data <- data.frame(z = as.numeric(z_grid), f = as.numeric(f_interp(z_grid)))

# 2. Create the plot
p <- ggplot(plot_data, aes(x = z, y = f)) +
  # Draw the interpolated smooth line from the line_data frame
  geom_line(data = line_data, color = "#8E44AD", linewidth = 1, alpha = 0.8) +
  # Draw the actual data points from the plot_data frame (inherited)
  geom_point(fill = "#D35400", color = "white", shape = 21, size = 3, stroke = 0.5) +
  # Labels and Title
  labs(
    title = "Marginal Posterior Density Estimation (beta[2,2])",
    subtitle = "Calculated via Vegas Integration with Spline Interpolation",
    x = "Parameter Space (z)",
    y = "Density",
    caption = "Source: Vegas R Integration"
  ) +
  # Apply a professional theme
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30", size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 20, 20, 20)
  )


print(p)

```


### Plot and compare with Stan

```{r}
plot(density(extract(stan_fit_2,"beta[2,2]")$"beta[2,2]"),main="beta[2,2]",col="blue",lwd=2)
points(myz,as.matrix(f_z),col="brown",pch=21,bg="brown")
f_interp <- splinefun(myz, as.matrix(f_z), method = "fmm")
lines(myz,f_interp(myz),col="magenta")
```
  
