---
title: "Bayesian Posterior Densities using Vegas"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Bayesian Posterior Densities using Vegas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=TRUE
)

# Check if running on GitHub Actions
# This is important as otherwise Github will attempt to build this and it's missing the python libraries and will fail
on_github <- Sys.getenv("GITHUB_ACTIONS") == "true"

   
# If on GitHub, set eval = FALSE for all subsequent chunks
knitr::opts_chunk$set(eval = !on_github)

knitr::opts_chunk$set(eval = FALSE)
```

```{r cran_check_setting,include=FALSE}
# this is to de-tune all examples to get past cran checks
is_checking <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))
if (is_checking) {
     n_cores <- 2 # number of cores for computing marginal density on cran
     no_den_pts<-5 # very ligth grid for on cran
    } else { # running locally to go higher
    n_cores <- parallel::detectCores() - 1
    no_den_pts=20
}

```

## Quickstart
This is a first example of how to compute a Bayesian posterior density using vegasr, with the integrand written as an R function. Examples of how to use vegasr to compute general integrals can be found in the ?vegas help page. See the package vignettes for other integrands and how to use Rcpp to write the integrand functions.

## Model Formulation

The model implemented here is a hierarchical Bayesian model with logistic link function and single predictor denoting treatment effect:

$$
\begin{aligned}
\mu_0 &\sim \text{Normal}(0, 2.5)\\
\sigma_0 &\sim \text{Half-Normal}(0, 2.5) \\
\mu_1 &\sim \text{Normal}(0, 2.5) \\
\sigma_1 &\sim \text{Half-Normal}(0, 2.5) \\
\beta_0 &\sim \text{Normal}(\mu_0, \sigma_0) \\
\beta_1 &\sim \text{Normal}(\mu_1, \sigma_1) \\
\text{logit(}{p_i}) &= \beta_0 + \beta_1 z_i\quad\quad\quad\quad\enspace\enspace\text{for}\enspace{i=1,\dots,N}  \\
y_i &\sim \text{Bernoulli}(p_i)
\end{aligned}
$$

## Data set

The R chunk below creates a simple dataset of three cols:

-   a binary response variable y (0/1 = non-responder/responder),
-   a (dummy) basket ID variable (=1)
-   a binary treatment variable (0/1 = control/test treatment)

The basket ID is currently set fixed at 1, denoting there is only one basket in this trial, i.e. it's a classical two arm randomized trial design. Other examples in the package include five baskets.

```{r datasets, include=TRUE}
thedata<-vegasr:::fn_create_data_1(99999) # a list of y and treat with both as matrices
```


## Construct Log Posterior Function 

To use direct integration for Bayesian computation the key task is to:

-  write a function that computes the log likelihood + log priors + log Jacobian

We recommend using a **change of variables** to deal with the ranges of integration going to $\infty$ or $-\infty$. This means we integrate over transformed variables (see below) which have finite bounds, typically $-1$, $0$ or $+1$. As we are changing variables then this requires the usual Jacobian adjustment to the function being integrated to ensure the volume being computed is equivalent to that on the original scale.

To map a variable $t\in(-1,1)$ to $x\in(-\infty,\infty)$ we can use the following function and its Jacobian (J):
$$
\begin{aligned}
x&=&\frac{t}{1-t^2} \\
\\
\enspace\text{where } \quad\enspace J &=& \frac{1+t^2}{(1-t^2)^2}
\end{aligned}
$$

## Find Log Evidence

The function below `vegasr:::fn_log_post_1` returns the log posterior. See ?fn_log_post_1 for specifics. The function `vegasBayesEvidence()` computes the log evidence, i.e. it integrates out all parameters in the posterior `vegasr:::fn_log_post_1` to give the constant needed to standardize the posterior to unity. Details of the parameters can be found in ?vegasBayesEvidence. 

```{r logpos,include=TRUE}
### Now use VEGAS
library(vegasr)
# now setup python environment
vegas_initialize() # this needs called once per session after library(vegasr) 

result_logEv<-vegasBayesEvidence(f=vegasr:::fn_log_post_1,
              lower=c(-1,-1,-1,-1,0.0001,0.0001), 
              upper=c(1,1,1,1,1,1),
              nitn_warm = 10, neval_warm = 10000,
              nitn = 10, neval = 10000,
              errTol=1,maxIter=10,seed=99999,nsearch=10000,
              extra_args=list(
                y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.))
cat("log evidence = ",result_logEv,"\n")



```


## Compute the Marginal Density

We use `vegasBayesPosterior()` to compute a single density value on the marginal posterior density for a single model parameter. This function takes as input a function for log posterior for the model but where one dimension is now fixed (here $\beta_0$ in the model above), i.e. the integrand has one less dimension, e.g. `vegasr:::fn_log_post_1` is 6-D whereas `vegasr:::fn_marg_1_1` below is 5-D and the 6th dimension is passed as a fixed value, z.  

```{r marg_single,include=TRUE}
# compute posterior marginal for intercept $\beta_0$ in the model where intercept = -1.
mymarg<-vegasBayesPosterior(f=vegasr:::fn_marg_1,
                           lower=c(-1,-1,-1,0.0001,0.0001),
                           upper=c(1,1,1,1,1),
                           nitn_warm = 10, neval_warm = 10000,
                           nitn = 10, neval = 10000,
                           errTol=1,maxIter=10,seed=99999,nsearch=10000,
                           log_evidence = result_logEv,
                           extra_args=list(
                             y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.,z=-1.))
cat("Marginal density f(z) at z = -1. = ",mymarg,"\n")



```

## Marginal Density over a Range
To compute the marginal posterior density over a range of z we iterate across a grid of z values. This can easily be done using doParallel as this is embarrassingly parallel - the computation of the posterior density at each fixed value of z are independent. 

```{r marg_vector,include=TRUE}

if(!nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))){
library(foreach)
library(doParallel)
library(extraDistr)
library(tictoc)

cl <- makeCluster(parallel::detectCores() - 1)

registerDoParallel(cl)
myz<-seq(-2.5,-0.,len=no_den_pts)
tic("Parallel Vegas Loop") # Start timer with a label
f_z<-foreach(z= myz,.packages = c("extraDistr", "vegasr")) %dopar% {
 vegasBayesPosterior(f=vegasr:::fn_marg_1,
                      lower=c(-1,-1,-1,0.0001,0.0001),
                      upper=c(1,1,1,1,1),
                      nitn_warm = 10, neval_warm = 10000,
                      nitn = 10, neval = 10000,
                      errTol=1,maxIter=10,seed=99999,nsearch=10000,
                      log_evidence = result_logEv,
                      extra_args=list(
                        y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.,z=z))

 }

stopCluster(cl)
toc() # Stops timer and prints
}
```

```{r plot_marg, include=FALSE,  fig.width=10, fig.height=6, out.width="100%", fig.align='center'}
if(!nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))){
library(ggplot2)

# Create a smooth interpolation for the line
z_grid <- seq(min(myz), max(myz), length.out = 500)
f_interp <- splinefun(myz, f_z, method = "fmm")

# Prepare data frames
plot_data <- data.frame(z = as.numeric(myz), f = as.numeric(f_z))
line_data <- data.frame(z = as.numeric(z_grid), f = as.numeric(f_interp(z_grid)))

# 2. Create the plot
p <- ggplot(plot_data, aes(x = z, y = f)) +
  # Draw the interpolated smooth line from the line_data frame
  geom_line(data = line_data, color = "#8E44AD", linewidth = 1, alpha = 0.8) +
  # Draw the actual data points from the plot_data frame (inherited)
  geom_point(fill = "#D35400", color = "white", shape = 21, size = 3, stroke = 0.5) +
  # Labels and Title
  labs(
    title = "Marginal Posterior Density Estimation",
    subtitle = "Calculated via Vegas Integration with Spline Interpolation",
    x = "Parameter Space (z)",
    y = "Density",
    caption = "Source: Vegas R Integration"
  ) +
  # Apply a professional theme
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30", size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 20, 20, 20)
  )

}
```

```{r plot_marg2, include=TRUE, fig.width=10, fig.height=6, out.width="100%", fig.align='center'}
# 3. Display the plot
# plotting code above in hidden chunk
if(!nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))){
print(p)
}
```
