---
title: "Bayesian Logistic Regression"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Bayesian Logistic Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Check if running on GitHub Actions
# This is important as otherwise Github will attempt to build this and it's missing the python libraries and will fail
#on_github <- Sys.getenv("GITHUB_ACTIONS") == "true"

# If on GitHub, set eval = FALSE for all subsequent chunks
#knitr::opts_chunk$set(eval = !on_github)

```

### Quickstart
tbc

## Model Formulation

The model implemented here is a non-centered parameterization (helps improve MCMC sampling) for a logistic model.

$$
\begin{aligned}
\mu_0 &\sim \text{Normal}(0, 2.5)\\
\sigma_0 &\sim \text{Half-Normal}(0, 2.5) \\
\mu_1 &\sim \text{Normal}(0, 2.5) \\
\sigma_1 &\sim \text{Half-Normal}(0, 2.5) \\
\tilde{\theta}_j &\sim \text{Normal}(0, 1)\quad\enspace\text{for}\enspace{j=1,2}  \\
\beta_0 &= \mu_0 + \sigma_0 \tilde{\theta}_1 \\
\beta_1 &= \mu_1 + \sigma_1 \tilde{\theta}_2 \\
\text{logit(}{p_i}) &= \beta_0 + \beta_1 z_i\quad\quad\enspace\enspace\text{for}\enspace{i=1,\dots,N}  \\
y_i &\sim \text{Bernoulli}(p_i)
\end{aligned}
$$
When using direct integration via Vegas we don't need the non-centered parameterization:

$$
\begin{aligned}
\beta_0 &\sim \text{Normal}(\mu_0, \sigma_0) \\
\beta_1 &\sim \text{Normal}(\mu_1, \sigma_1) \\
\end{aligned}
$$
## Data set

The R chunk below creates a simple dataset of three cols:

-   a binary response variable y (0/1 = non-responder/responder),
-   a (dummy) basket ID variable (=1)
-   a binary treatment variable (0/1 = control/test treatment)

The basket ID is currently set fixed at 1, denoting there is only one basket in this trial, i.e. a classical two arm randomized trial design. Baskets will be added in other vignettes.

```{r datasets, include=TRUE,eval=TRUE}
### Prepare model inputs
set.seed(99999)
# Set up data
rr_k_ctrl <- c(0.20)        # control response rate for each basket
rr_k_trt <- c(0.40)         # treatment response rate for each basket

K<-length(rr_k_ctrl)        # number of baskets

N_k_ctrl <- rep(100, K)     # number of control participants per basket
N_k_trt <- rep(100, K)      # number of treatment participants per basket
N_k <- N_k_ctrl + N_k_trt   # number of participants per basket (both arms combined)
N <- sum(N_k)               # total sample size
k_vec <- rep(1:K, N_k)      # N x 1 vector of basket indicators (1 to K)

z_vec<-NULL;
y<-NULL;
for(i in 1:K){ # for each basket repeat 0-control 1-trt according to the specifc Ns
  z_vec<-c(z_vec,rep(0:1,c(N_k_ctrl[i],N_k_trt[i]))) # treatment/control indicator
  y<-c(y,
       c(rbinom(N_k_ctrl[i],1,rr_k_ctrl[i]), # bernoulli for control
         rbinom(N_k_trt[i],1,rr_k_trt[i]))) #           for trt
}

thedata<-data.frame(y,basketID=k_vec,Treatment=z_vec)

```


## 1. Construct Log Posterior Function 

To use direct integration in Vegas the key parts are:

-  define the prior density functions
-  define the log likelihood + log priors

In addition, we also implement a **change of variables** to deal with the range of integration going to $\infty$. This means we integrate over transformed variables (see below) which have finite bounds. As we are changing variables then this requires the usual Jacobian adjustment to the function being integrated to ensure the volume being computed is equivalent to that on the original scale.

To map a variable $t\in(-1,1)$ to $x\in(-\infty,\infty)$ we can use the following function and it's Jacobian (J):
$$
\begin{aligned}
x&=&\frac{t}{1-t^2} \\
\\
\enspace\text{where } \quad\enspace J &=& \frac{1+t^2}{(1-t^2)^2}
\end{aligned}
$$

The function below `compute_log_lik(theta, y,T)` is necessary to compute the overall standardization constant - the constant which ensures the posterior density integrates to unity - and the integration here is over all parameters in the model. We work in log scale as much as possible and a scaling ``trick'' is used to help avoid numerical overflow. These technical numerical details can be found in the full quarto file.  

```{r logpos,include=TRUE,eval=TRUE}
library(extraDistr)

# response
y<-matrix(data=as.numeric(thedata$y),ncol=1)
# treatment
treat<-matrix(data=as.numeric(thedata$Treatment),ncol=1)
# matrix of samole parameter values - nrow is batch, ncol is model dimension
dummytheta<-matrix(data=c(-0.11,-0.13,0.15,0.11,0.051,0.052,
                          -0.12,-0.1,0.17,0.12,0.052,0.051,
                          -0.13,-0.11,0.11,0.19,0.053,0.054
                          ),ncol=6,byrow=TRUE)

## Define log posterior inclding change of variables
r_compute_log_lik<-function(theta,   # matrix Batch x M
                            y,       # matrix N x 1
                            treat,   # matrix N x 1
                            scaling, # scalar
                            s        # scalar  
                            ){

  # this is a trimming function to avoid extreme end of integral limits, 
  # same as -inf,+inf 
  theta0=pmin(pmax(theta[,1], -0.9999), 0.9999)
  theta1=pmin(pmax(theta[,2], -0.9999), 0.9999)
  theta2=pmin(pmax(theta[,3], -0.9999), 0.9999)
  theta3=pmin(pmax(theta[,4], -0.9999), 0.9999)
  theta4=pmin(pmax(theta[,5], 0.0001), 0.9999)
  theta5=pmin(pmax(theta[,6], 0.0001), 0.9999)

  jacobianL = (
    log1p(theta0^2) - 2.0*log1p(-(theta0^2))
    + log1p(theta1^2) - 2.0*log1p(-(theta1^2))
    + log1p(theta2^2) - 2.0*log1p(-(theta2^2))
    + log1p(theta3^2) - 2.0*log1p(-(theta3^2))
    + log1p(theta4^2) - 2.0*log1p(-(theta4^2))
    + log1p(theta5^2) - 2.0*log1p(-(theta5^2))

  )

  a0=theta0/(1-theta0^2)
  a1=theta1/(1-theta1^2)
  mu0=theta2/(1-theta2^2)
  mu1=theta3/(1-theta3^2)
  sigma0=theta4/(1-theta4^2)
  sigma1=theta5/(1-theta5^2)

  # eta=a0+a1*treat # (10,3) where a0 and a1 = (3,) and treat is (10,1) 
  #                                                broadcasts to (10,3)
  # R doesn't have auto broadcast but this is equivalent
  eta <- sweep(treat %*% rbind(a1), 2, a0, "+")
  # logL = sum_i^n ( y_i*(a0+a1*T_i) - log(1+ exp(a0+a1*T_i) )  )
  logL = apply(sweep(eta,1,y,"*") - log1p(exp(eta)),2,sum)

  # now add the priors
  prior_a0 = dnorm(a0,mean=mu0,sd=sigma0,log=TRUE)
  prior_a1 = dnorm(a1,mean=mu1,sd=sigma1,log=TRUE)
  prior_mu0 = dnorm(mu0,mean=0.,sd=2.5,log=TRUE)
  prior_mu1 = dnorm(mu1,mean=0.,sd=2.5,log=TRUE)
  prior_sigma0 = dhnorm(sigma0, sigma=2.5,log=TRUE)
  prior_sigma1 = dhnorm(sigma1, sigma=2.5,log=TRUE)

  logDens = logL + prior_a0 + prior_a1 + prior_mu0 + prior_mu1 + 
                                      prior_sigma0 + prior_sigma1
  logPost = logDens + jacobianL

  if(s==1.){ # search phase for max - keep in log
           return(logPost - scaling[1])
  } else return(exp(logPost - scaling[1]) ) # integrand eval - use raw

    }

## test if log_posterior works - pass matrix of parameter values 
r_compute_log_lik(theta=dummytheta,y=y,treat=treat,scaling=0,s=1.)



```

## 2. Find Scaling Factor and Integrate 

To mitigate underflow we work in logs as much as possible and scaling can also assist here where all the log posterior values have a location shift so the maximum is close to zero on a log scale. This maximum need not be exact and we use a simple and fast grid search. 

We use vegas to integrate over the log posterior to get the log evidence, i.e. the standardization constant, that ensures the posterior integrates to unity. 

```{r logpos2,include=TRUE,eval=TRUE}
set.seed(11111)
start<-c(-1,-1,-1,-1,0.0001,0.0001)
stop<-c(1,1,1,1,1,1)
searchPts <- mapply(seq, from = start, to = stop, 
                    MoreArgs = list(length.out = 1000))
for(i in 1:6){searchPts[,i]<-searchPts[sample(1:nrow(searchPts)),i]}

mymax<-max(r_compute_log_lik(theta=searchPts,y=y,treat=treat,
                             scaling=0.,
                             s=1.))
# location shift - approx max over the integration domain
print(mymax)


### Now use VEGAS
library(vegasr)
# now setup python environment
vegas_initialize() # this needed called once per session after library(vegas) 

# perform numerical integration
result<-vegas_integrate(f=r_compute_log_lik,
                        lower=c(-1,-1,-1,-1,0.0001,0.0001), 
                        upper=c(1,1,1,1,1,1),
                        nitn_warm = 10, neval_warm = 10000,
                        nitn = 10, neval = 10000,
                        errTol=1,maxIter=10,
                        y=y,treat=treat,scaling=mymax,s=0.)
print(result)

log_evidence = mymax + log(result$mean)
cat("log evidence = ",log_evidence,"\n")
# should be around -129.4




```



