---
title: "Bayesian Logistic Regression"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Bayesian Logistic Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Check if running on GitHub Actions
# This is important as otherwise Github will attempt to build this and it's missing the python libraries and will fail
#on_github <- Sys.getenv("GITHUB_ACTIONS") == "true"

# If on GitHub, set eval = FALSE for all subsequent chunks
#knitr::opts_chunk$set(eval = !on_github)

```

```{r cran_check_setting,include=FALSE,eval=TRUE}
# this is to de-tune all examples to get past cran checks
is_checking <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))
if (is_checking) {
     n_cores <- 2 # number of cores for computing marginal density on cran
     no_den_pts<-5 # very ligth grid for on cran
    } else { # running locally to go higher
    n_cores <- parallel::detectCores() - 1
    no_den_pts=50
}

```

### Quickstart
tbc

## Model Formulation

The model implemented here is a non-centered parameterization (helps improve MCMC sampling) for a logistic model.

$$
\begin{aligned}
\mu_0 &\sim \text{Normal}(0, 2.5)\\
\sigma_0 &\sim \text{Half-Normal}(0, 2.5) \\
\mu_1 &\sim \text{Normal}(0, 2.5) \\
\sigma_1 &\sim \text{Half-Normal}(0, 2.5) \\
\tilde{\theta}_j &\sim \text{Normal}(0, 1)\quad\enspace\text{for}\enspace{j=1,2}  \\
\beta_0 &= \mu_0 + \sigma_0 \tilde{\theta}_1 \\
\beta_1 &= \mu_1 + \sigma_1 \tilde{\theta}_2 \\
\text{logit(}{p_i}) &= \beta_0 + \beta_1 z_i\quad\quad\enspace\enspace\text{for}\enspace{i=1,\dots,N}  \\
y_i &\sim \text{Bernoulli}(p_i)
\end{aligned}
$$
When using direct integration via Vegas we don't need the non-centered parameterization:

$$
\begin{aligned}
\beta_0 &\sim \text{Normal}(\mu_0, \sigma_0) \\
\beta_1 &\sim \text{Normal}(\mu_1, \sigma_1) \\
\end{aligned}
$$

### Data set

The R chunk below creates a simple dataset of three cols:

-   a binary response variable y (0/1 = non-responder/responder),
-   a (dummy) basket ID variable (=1)
-   a binary treatment variable (0/1 = control/test treatment)

The basket ID is currently set fixed at 1, denoting there is only one basket in this trial, i.e. a classical two arm randomized trial design. Baskets will be added in other vignettes.

```{r datasets, include=TRUE,eval=TRUE}
thedata<-vegasr:::fn_create_data_1(99999) # a list of y and treat as matrices
```


## 1. Construct Log Posterior Function 

To use direct integration in Vegas the key parts are:

-  define the prior density functions
-  define the log likelihood + log priors

In addition, we also implement a **change of variables** to deal with the range of integration going to $\infty$. This means we integrate over transformed variables (see below) which have finite bounds. As we are changing variables then this requires the usual Jacobian adjustment to the function being integrated to ensure the volume being computed is equivalent to that on the original scale.

To map a variable $t\in(-1,1)$ to $x\in(-\infty,\infty)$ we can use the following function and it's Jacobian (J):
$$
\begin{aligned}
x&=&\frac{t}{1-t^2} \\
\\
\enspace\text{where } \quad\enspace J &=& \frac{1+t^2}{(1-t^2)^2}
\end{aligned}
$$

The function below `compute_log_lik(theta, y,T)` is necessary to compute the overall standardization constant - the constant which ensures the posterior density integrates to unity - and the integration here is over all parameters in the model. We work in log scale as much as possible and a scaling ``trick'' is used to help avoid numerical overflow. These technical numerical details can be found in the full quarto file.  

```{r logpos,include=TRUE,eval=TRUE}
### Now use VEGAS
library(vegasr)
# now setup python environment
vegas_initialize() # this needed called once per session after library(vegas) 

result_logEv<-vegasBayesEvidence(f=vegasr:::fn_log_post_1,
              lower=c(-1,-1,-1,-1,0.0001,0.0001), 
              upper=c(1,1,1,1,1,1),
              nitn_warm = 10, neval_warm = 10000,
              nitn = 10, neval = 10000,
              errTol=1,maxIter=10,seed=99999,nsearch=10000,
              extra_args=list(
                y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.))
cat("log evidence = ",result_logEv,"\n")



```


## 2. Find Log Evidence

We use vegas to integrate over the log posterior to get the log evidence, i.e. the standardization constant, that ensures the posterior integrates to unity. Also known as the log marginal likelihood.

```{r marg_single,include=TRUE,eval=TRUE}

mymarg<-vegasBayesPosterior(f=vegasr:::fn_marg_1_1,
                           lower=c(-1,-1,-1,0.0001,0.0001),
                           upper=c(1,1,1,1,1),
                           nitn_warm = 10, neval_warm = 10000,
                           nitn = 10, neval = 10000,
                           errTol=1,maxIter=10,seed=99999,nsearch=10000,
                           log_evidence = result_logEv,
                           extra_args=list(
                             y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.,z=-1.))
cat("Marginal density f(z) at z = -1. = ",mymarg,"\n")



```


```{r marg_vector,include=TRUE,eval=TRUE}

library(foreach)
library(doParallel)
library(extraDistr)
library(tictoc)

cl <- makeCluster(n_cores)

registerDoParallel(cl)

myz<-seq(-2.5,-0.,len=no_den_pts)
tic("Parallel Vegas Loop") # Start timer with a label
f_z<-foreach(z= myz,.packages = c("extraDistr", "vegasr")) %dopar% {
 vegasBayesPosterior(f=vegasr:::fn_marg_1_1,
                      lower=c(-1,-1,-1,0.0001,0.0001),
                      upper=c(1,1,1,1,1),
                      nitn_warm = 10, neval_warm = 10000,
                      nitn = 10, neval = 10000,
                      errTol=1,maxIter=10,seed=99999,nsearch=10000,
                      log_evidence = result_logEv,
                      extra_args=list(
                        y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.,z=z))

 }

stopCluster(cl)
toc() # Stops timer and prints

```

```{r plot_marg, include=FALSE, eval=TRUE, fig.width=10, fig.height=6, out.width="100%", fig.align='center'}
#plot(myz,f_z,main="Vegas R",col="blue",lwd=2,type="n")
#points(myz,f_z,col="brown",pch=21,bg="brown")
#f_interp <- splinefun(myz, f_z, method = "fmm")
#lines(myz,f_interp(myz),col="magenta")

library(ggplot2)

# 1. Prepare data for plotting
# Ensure f_z is a numeric vector (in case it is a list from foreach)
if (is.list(f_z)) {
  # If it's a list of results, extract the 'mean' component
  if (is.list(f_z[[1]]) && "mean" %in% names(f_z[[1]])) {
    f_z_vec <- unlist(lapply(f_z, function(x) x$mean))
  } else {
    f_z_vec <- unlist(f_z)
  }
} else {
  f_z_vec <- f_z
}

# Create a smooth interpolation for the line
z_grid <- seq(min(myz), max(myz), length.out = 500)
f_interp <- splinefun(myz, f_z_vec, method = "fmm")

# Prepare data frames
plot_data <- data.frame(z = as.numeric(myz), f = as.numeric(f_z_vec))
line_data <- data.frame(z = as.numeric(z_grid), f = as.numeric(f_interp(z_grid)))

# 2. Create the plot
p <- ggplot(plot_data, aes(x = z, y = f)) +
  # Draw the interpolated smooth line from the line_data frame
  geom_line(data = line_data, color = "#8E44AD", linewidth = 1, alpha = 0.8) +
  # Draw the actual data points from the plot_data frame (inherited)
  geom_point(fill = "#D35400", color = "white", shape = 21, size = 3, stroke = 0.5) +
  # Labels and Title
  labs(
    title = "Marginal Posterior Density Estimation",
    subtitle = "Calculated via Vegas Integration with Spline Interpolation",
    x = "Parameter Space (z)",
    y = "Density",
    caption = "Source: Vegas R Integration"
  ) +
  # Apply a professional theme
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30", size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 20, 20, 20)
  )

```

```{r plot_marg2, include=TRUE, eval=TRUE, fig.width=10, fig.height=6, out.width="100%", fig.align='center'}
# 3. Display the plot
# plotting code above in hidden chunk
print(p)

```

