---
title: "Using Rcpp in Vegasr"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Using Rcpp in Vegasr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=TRUE
)

# options(pkgdown.parallel_vignettes = FALSE)
# Check if running on GitHub Actions
# This is important as otherwise Github will attempt to build this and it's missing the python libraries and will fail
on_github <- Sys.getenv("GITHUB_ACTIONS") == "true"

   
# If on GitHub, set eval = FALSE for all subsequent chunks
#knitr::opts_chunk$set(eval = !on_github)

knitr::opts_chunk$set(eval=TRUE)

```

```{r cran_check_setting,include=FALSE}
# this is to de-tune all examples to get past cran checks
is_checking <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))
if (is_checking) {
     n_cores <- 2 # number of cores for computing marginal density on cran
     no_den_pts<-5 # very ligth grid for on cran
    } else { # running locally to go higher
    n_cores <- parallel::detectCores() - 1
no_den_pts=20
}

```

## Quickstart
Using Rcpp libraries to write the log posterior function will almost surely be faster, perhaps many times faster, than an R function even when vectorized. Examples of the timing comparison are below.   

Example log posterior functions are given below using two Rcpp linear algebra libraries, [RcppArmadillo](https://github.com/rcppcore/rcpparmadillo) and [RcppEigen](https://github.com/RcppCore/RcppEigen). These libraries have different pros and cons. RcppArmadillo has some built-in parallelization (via OpenMP - not supported on MacOS) while RcppEigen does not, but both are highly efficient in their own way. RcppEigen works well with [RcppParallel](https://github.com/rcppcore/rcppparallel) which does not use OpenMP and can be more problematic when used with RcppArmadillo. 

## Model Formulation

The model implemented here is:

$$
\begin{aligned}
\mu_0 &\sim \text{Normal}(0, 2.5)\\
\sigma_0 &\sim \text{Half-Normal}(0, 2.5) \\
\mu_1 &\sim \text{Normal}(0, 2.5) \\
\sigma_1 &\sim \text{Half-Normal}(0, 2.5) \\
\beta_0 &\sim \text{Normal}(\mu_0, \sigma_0) \\
\beta_1 &\sim \text{Normal}(\mu_1, \sigma_1) \\
\text{logit(}{p_i}) &= \beta_0 + \beta_1 z_i\quad\quad\quad\quad\enspace\enspace\text{for}\enspace{i=1,\dots,N}  \\
y_i &\sim \text{Bernoulli}(p_i)
\end{aligned}
$$


## Data set
The R chunk below creates a simple dataset of three cols:

-   a binary response variable y (0/1 = non-responder/responder),
-   a (dummy) basket ID variable (=1)
-   a binary treatment variable (0/1 = control/test treatment)

The basket ID is currently set fixed at 1, denoting there is only one basket in this trial, i.e. a classical two arm randomized trial design.

```{r datasets, include=TRUE}
library(vegasr)
vegas_initialize()
thedata<-vegasr:::fn_create_data_1(99999) # a list of y and treat as matrices
# this function is in vegasr/R/fn_internal.R
```
## Example of calling log posterior functions
These functions are available in the source package in fns_arma.cpp or fns_eigen.cpp in /src. They implement the same log posterior integrand, using different libraries, and the numerical output values should be the same.

```{r logpos,include=TRUE}
### Call the Rcpp functions to demonstrate correct input and output values

theta<- matrix(data=rep(0.1,length=4*6), ncol = 6) # dummy data - must be 6 cols, any number of rows

vegasr:::fn_log_post_1(theta, thedata$y, thedata$treat,.0, 1.0) # R function
# this function is in vegasr/R/fn_internal.R

vegasr::arma_fn_log_post_1(theta, thedata$y, thedata$treat,0.0, 1.0) # Armadillo function
# this function is in vegasr/src/fns_arma.cpp

vegasr::eigen_fn_log_post_1(theta, thedata$y, thedata$treat,0.0, 1.0) # Eigen function
# this function is in vegasr/src/fns_eigen.cpp

```
## Timing comparisons
Results below demonstrate the speed increase of Rcpp over plain R. An approx x4 speed up is seen in this example on a like for like basis. Using parallel computation additionally makes a big difference even in this fairly simple case, with a bigger differential expected the more complex the model.  When computing marginal densities and/or integrands in more than a few dimensions the speed of each individual function call becomes increasing important as this will be called a large number of times. 

```{r logpos2,include=TRUE}
library(vegasr)
# now setup python environment
vegas_initialize() # this needed called once per session after library(vegas)
library(tictoc) # easy timing

## pure R
tic("plain R")
result_logEv<-vegasBayesEvidence(f=vegasr:::fn_log_post_1,
                                 lower=c(-1,-1,-1,-1,0.0001,0.0001),
                                 upper=c(1,1,1,1,1,1),
                                 nitn_warm = 5, neval_warm = 1e5,
                                 nitn = 5, neval = 1e5,
                                 errTol=1,maxIter=10,seed=99999,nsearch=10000,
                                 extra_args=list(
                                   y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.))
toc()
cat("log evidence = ",result_logEv,"\n\n")

## Armadillo
tic("Armadillo")
result_logEv<-vegasBayesEvidence(f=vegasr:::arma_fn_log_post_1,
                                 lower=c(-1,-1,-1,-1,0.0001,0.0001),
                                 upper=c(1,1,1,1,1,1),
                                 nitn_warm = 5, neval_warm = 1e5,
                                 nitn = 5, neval = 1e5,
                                 errTol=1,maxIter=10,seed=99999,nsearch=10000,
                                 extra_args=list(
                                   y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.))
toc()
cat("log evidence = ",result_logEv,"\n\n")

## Eigen
tic("Eigen")
result_logEv<-vegasBayesEvidence(f=vegasr::eigen_fn_log_post_1,
                                 lower=c(-1,-1,-1,-1,0.0001,0.0001),
                                 upper=c(1,1,1,1,1,1),
                                 nitn_warm = 5, neval_warm = 1e5,
                                 nitn = 5, neval = 1e5,
                                 errTol=1,maxIter=10,seed=99999,nsearch=10000,
                                 extra_args=list(
                                   y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.))
toc()
cat("log evidence = ",result_logEv,"\n\n")
result_logEv_keep<-result_logEv;

## Eigen with parallel
tic("Eigen and RcppParallel")
result_logEv<-vegasBayesEvidence(f=vegasr::eigen_fn_log_post_1_par,
                                 lower=c(-1,-1,-1,-1,0.0001,0.0001),
                                 upper=c(1,1,1,1,1,1),
                                 nitn_warm = 5, neval_warm = 1e5, #this is deliberately too small
                                 nitn = 5, neval = 1e5, #this is deliberately too small
                                 errTol=1,maxIter=10,seed=99999,nsearch=10000,
                                 extra_args=list(
                                   y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.))
toc()
cat("log evidence = ",result_logEv,"\n\n")


```
## Compute Marginal Posterior Density
An example using a parallel version of log posterior via Eigen and RcppParallel libraries. See fns_eigen.cpp in /src in source package. 

```{r marg_single,include=TRUE}
# Marginal density at a single point
tic()
mymarg<-vegasBayesPosterior(f=vegasr::eigen_fn_marg_1,
                            lower=c(-1,-1,-1,0.0001,0.0001),
                            upper=c(1,1,1,1,1),
                            nitn_warm = 10, neval_warm = 10000,
                            nitn = 10, neval = 10000,
                            errTol=1,maxIter=10,seed=99999,nsearch=10000,
                            log_evidence = result_logEv,
                            extra_args=list(
                              y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.,z=-1.))

cat("Marginal density f(z) at z = -1. = ",mymarg,"\n")
toc()

## Recompute Log Evidence - we did this above, but this is to show issues of accuracy
tic()
result_logEv<-vegasBayesEvidence(f=vegasr::eigen_fn_log_post_1_par,
                                 lower=c(-1,-1,-1,-1,0.0001,0.0001),
                                 upper=c(1,1,1,1,1,1),
                                 nitn_warm = 5, neval_warm = 1e3, #this is too small
                                 nitn = 5, neval = 1e3, #this is too small
                                 errTol=1,maxIter=10,seed=99999,nsearch=10000,
                                 extra_args=list(
                                   y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.))
toc()
cat("log evidence = ",result_logEv,"\n\n")

##################################
# compute marginal across a range
myz<-seq(-2.5,-0.,len=no_den_pts)
tic("") # Start timer with a label
f_z<-rep(0,length(myz));
i<-1;
for(z in myz){
  f_z[i]<-vegasBayesPosterior(f=vegasr::eigen_fn_marg_1,
                            lower=c(-1,-1,-1,0.0001,0.0001),
                            upper=c(1,1,1,1,1),
                            nitn_warm = 10, neval_warm = 10000,
                            nitn = 10, neval = 10000,
                            errTol=1,maxIter=10,seed=99999,nsearch=10000,
                            log_evidence = result_logEv,
                            extra_args=list(
                              y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.,z=z))
  i<-i+1
}

toc() # Stops timer and prints

```

### Splines
While it is possible to compute a fine grid of density values for the marginal this is typically wasteful and highly computationally intensive. An alternative option is to compute a small set of accurate density values and then apply spline interpolation - which will go through each point - to give the complete density. This can then be integrated to get necessary quantiles or other statistics of interest. 

### Standardized using Log Evidence
The textbook way to compute the marginal density is to divide the density values by the evidence, typically kept on log scales as long as possible for numerical accuracy. We calculated the log_evidence above. In practice this can sometimes result in poorly standardized densities, i.e. do not integrate to unity, unless the log evidence is particularly accurate (as small differences on log scale get magnified when scaling the density) which can take a lot of computational effort. An alternative is to standardize the marginal explicitly via integrating it and dividing by its area, which is easy as its one-dimensional. This approach is shown below. 

Note that if the objective is not to estimate the full density but rather estimation of a quantile, e.g., 2.5% point, then it may be more efficient to compute the log evidence and then only a small set of density points in the relevant tail of the marginal. See next section.  

```{r marg_beta111,include=TRUE}
# explicitly standardize the posterior marginal
logun<-log(f_z)+result_logEv # revert back to unstandardizesd log density
f_z_un<-exp(logun) # back to unstand density
ff_interp <- splinefun(myz, as.matrix(f_z_un), method = "fmm") # fit fun to every point
evA<-integrate(ff_interp,min(myz),max(myz))$value # find area 
ff_z<-f_z_un/evA # divide by area to standardize marginal density
ff_interp <- splinefun(myz, as.matrix(ff_z), method = "fmm") # now

```


```{r plot_marg, include=FALSE, fig.width=10, fig.height=6, out.width="100%", fig.align='center'}

if(!nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))){
library(ggplot2)

# Create a smooth interpolation for the line
z_grid <- seq(min(myz), max(myz), length.out = 500)
f_interp <- splinefun(myz, f_z, method = "fmm")

# Prepare data frames
plot_data <- data.frame(z = as.numeric(myz), f = as.numeric(f_z))
line_data <- data.frame(z = as.numeric(z_grid), f = as.numeric(f_interp(z_grid)))

plot2_data <- data.frame(z = as.numeric(myz), f = as.numeric(ff_z))
line2_data <- data.frame(z = as.numeric(z_grid), f = as.numeric(ff_interp(z_grid)))

# 2. Create the plot
p <- ggplot() +
  # --- SECOND DATASET (Original Log Ev) ---
  geom_area(data = line_data, aes(x = z, y = f, fill = "Standardized via Log Evidence"), alpha = 0.1) +
  geom_line(data = line_data, aes(x = z, y = f, color = "Standardized via Log Evidence"), linewidth = 1, alpha = 0.6) +
  geom_point(data = plot_data, aes(x = z, y = f, fill = "Standardized via Log Evidence"), color = "white", shape = 21, size = 3, stroke = 0.5) +

  # --- FIRST DATASET (Marginal Posterior) ---
  geom_area(data = line2_data, aes(x = z, y = f, fill = "Standardized explicitly"), alpha = 0.2) +
  geom_line(data = line2_data, aes(x = z, y = f, color = "Standardized explicitly"), linewidth = 1, alpha = 0.8) +
  geom_point(data = plot2_data, aes(x = z, y = f, fill = "Standardized explicitly"), color = "white", shape = 21, size = 3, stroke = 0.5) +
  
  # Unified Legend
  scale_color_manual(name = "", values = c(
    "Standardized via Log Evidence" = "#8E44AD",
    "Standardized explicitly" = "#2E86C1"
  )) +
  scale_fill_manual(name = "", values = c(
    "Standardized via Log Evidence" = "#8E44AD",
    "Standardized explicitly" = "#2E86C1"
  )) +

  # Labels and Title
  labs(
    title = "Marginal Posterior Density Estimation",
    subtitle = "Calculated via Vegas Integration with Spline Interpolation",
    x = "Parameter Space (z)",
    y = "Density",
    caption = "Source: Vegas R Integration"
  ) +
  # Apply a professional theme
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30", size = 11),
    axis.title = element_text(face = "bold"),
    legend.position = "top",
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 20, 20, 20)
  )
}
```

```{r plot_marg2, include=TRUE, fig.width=10, fig.height=6, out.width="100%", fig.align='center'}
# 3. Display the plot
# plotting code above in hidden chunk
if(!nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))){
print(p)
}
```

### Computed Tail Probabilities
Often we are not interested in the full marginal posterior but we do want its tails, e.g. for 95% CI limits. Below shows simple example of how to do this. We compute the marginal density for a small number of points in the tail and as we known the log evidence we can use uniroot to solve the necessary quantile. 

```{r quantile,include=TRUE}
##################################
# compute marginal across a range for just the tail
myz<-seq(-2.5,-1.6,len=10)
tic("") # Start timer with a label
f_z<-rep(0,length(myz));
i<-1;
for(z in myz){
  f_z[i]<-vegasBayesPosterior(f=vegasr::eigen_fn_marg_1,
                            lower=c(-1,-1,-1,0.0001,0.0001),
                            upper=c(1,1,1,1,1),
                            nitn_warm = 10, neval_warm = 10000,
                            nitn = 10, neval = 10000,
                            errTol=1,maxIter=10,seed=99999,nsearch=10000,
                            log_evidence = result_logEv_keep,
                            extra_args=list(
                              y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.,z=z))
  cat("i=",i," z=",z," fz=",f_z[i],"\n")
  i<-i+1
}

f_interp <- splinefun(myz, f_z, method = "fmm")
myF<-function(x,target){return(integrate(f_interp,min(myz),x)$value-target)}

lower25<-uniroot(f=myF,interval=c(-2.5,-1.6),target=0.025)$root
cat("\n\nLower 2.5% of Marginal Posterior = ", lower25,"\n")
  
```




