---
title: "Using Rcpp in Vegasr"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Using Rcpp in Vegasr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Check if running on GitHub Actions
# This is important as otherwise Github will attempt to build this and it's missing the python libraries and will fail
on_github <- Sys.getenv("GITHUB_ACTIONS") == "true"

   
# If on GitHub, set eval = FALSE for all subsequent chunks
knitr::opts_chunk$set(eval = !on_github)

```

```{r cran_check_setting,include=FALSE,eval=TRUE}
# this is to de-tune all examples to get past cran checks
is_checking <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))
if (is_checking) {
     n_cores <- 2 # number of cores for computing marginal density on cran
     no_den_pts<-5 # very ligth grid for on cran
    } else { # running locally to go higher
    n_cores <- parallel::detectCores() - 1
    no_den_pts=50
}

```

### Quickstart
13th-May-update: Using Rcpp to write the log posterior function will almost surely be faster, perhaps many times faster, than an R function even when vectorized. Examples of the timing comparison are below. 

Example log posterior functions are given below using two Rcpp linear algebra libraries, [RcppArmadillo](https://github.com/rcppcore/rcpparmadillo) and [RcppEigen](https://github.com/RcppCore/RcppEigen). These libraries have different pros and cons. In short RcppArmadillo has some built-in parallelization while RcppEigen does not, but both are highly efficient in their own way. RcppEigen works well with [RcppParallel](https://github.com/rcppcore/rcppparallel) which can be more problematic with RcppArmadillo. 

## Model Formulation


### Data set

```{r datasets, include=TRUE,eval=TRUE}
library(vegasr)
vegas_initialize()
thedata<-vegasr:::fn_create_data_1(99999) # a list of y and treat as matrices
# this function is in vegasr/R/fn_internal.R
```

```{r logpos,include=TRUE,eval=TRUE}
### Now use VEGAS

theta<- matrix(data=rep(0.1,length=4*6), ncol = 6)

vegasr:::fn_log_post_1(theta, thedata$y, thedata$treat,.0, 1.0)
# this function is in vegasr/R/fn_internal.R

vegasr::arma_fn_log_post_1(theta, thedata$y, thedata$treat,0.0, 1.0)
# this function is in vegasr/src/fns_arma.cpp

vegasr::eigen_fn_log_post_1(theta, thedata$y, thedata$treat,0.0, 1.0)
# this function is in vegasr/src/fns_eigen.cpp

```
```{r logpos2,include=TRUE,eval=TRUE}
library(vegasr)
# now setup python environment
vegas_initialize() # this needed called once per session after library(vegas)
library(tictoc)


tic()
result_logEv<-vegasBayesEvidence(f=vegasr:::fn_log_post_1,
                                 lower=c(-1,-1,-1,-1,0.0001,0.0001),
                                 upper=c(1,1,1,1,1,1),
                                 nitn_warm = 5, neval_warm = 1e5,
                                 nitn = 5, neval = 1e5,
                                 errTol=0.1,maxIter=10,seed=99999,nsearch=10000,
                                 extra_args=list(
                                   y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.))
toc()
cat("log evidence = ",result_logEv,"\n")


tic()
result_logEv<-vegasBayesEvidence(f=vegasr::arma_fn_log_post_1,
                                 lower=c(-1,-1,-1,-1,0.0001,0.0001),
                                 upper=c(1,1,1,1,1,1),
                                 nitn_warm = 5, neval_warm = 1e5,
                                 nitn = 5, neval = 1e5,
                                 errTol=0.1,maxIter=10,seed=99999,nsearch=10000,
                                 extra_args=list(
                                   y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.))
toc()
cat("log evidence = ",result_logEv,"\n")

tic()
result_logEv<-vegasBayesEvidence(f=vegasr::eigen_fn_log_post_1,
                                 lower=c(-1,-1,-1,-1,0.0001,0.0001),
                                 upper=c(1,1,1,1,1,1),
                                 nitn_warm = 5, neval_warm = 1e5,
                                 nitn = 5, neval = 1e5,
                                 errTol=0.1,maxIter=10,seed=99999,nsearch=10000,
                                 extra_args=list(
                                   y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.))
toc()
cat("log evidence = ",result_logEv,"\n")

tic()
result_logEv<-vegasBayesEvidence(f=vegasr::eigen_fn_log_post_1_par,
                                 lower=c(-1,-1,-1,-1,0.0001,0.0001),
                                 upper=c(1,1,1,1,1,1),
                                 nitn_warm = 5, neval_warm = 1e5,
                                 nitn = 5, neval = 1e5,
                                 errTol=0.1,maxIter=10,seed=99999,nsearch=10000,
                                 extra_args=list(
                                   y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.))
toc()
cat("log evidence = ",result_logEv,"\n")


```
## 2. Find Marginal

tbc

```{r marg_single,include=TRUE,eval=TRUE}

tic()
mymarg<-vegasBayesPosterior(f=vegasr::eigen_fn_marg_1_1_par,
                            lower=c(-1,-1,-1,0.0001,0.0001),
                            upper=c(1,1,1,1,1),
                            nitn_warm = 10, neval_warm = 10000,
                            nitn = 10, neval = 10000,
                            errTol=1,maxIter=10,seed=99999,nsearch=10000,
                            log_evidence = result_logEv,
                            extra_args=list(
                              y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.,z=-1.))
cat("Marginal density f(z) at z = -1. = ",mymarg,"\n")
toc()

myz<-seq(-2.5,-0.,len=no_den_pts)
tic("") # Start timer with a label
f_z<-rep(0,length(myz));
i<-1;
for(z in myz){
  f_z[i]<-vegasBayesPosterior(f=vegasr::eigen_fn_marg_1_1_par,
                            lower=c(-1,-1,-1,0.0001,0.0001),
                            upper=c(1,1,1,1,1),
                            nitn_warm = 10, neval_warm = 10000,
                            nitn = 10, neval = 10000,
                            errTol=1,maxIter=10,seed=99999,nsearch=10000,
                            log_evidence = result_logEv,
                            extra_args=list(
                              y=thedata$y,treat=thedata$treat,shiftby=0,uselog=1.,z=z))
  i<-i+1
}

toc() # Stops timer and prints

```
```{r plot_marg, include=FALSE, eval=TRUE, fig.width=10, fig.height=6, out.width="100%", fig.align='center'}
#plot(myz,f_z,main="Vegas R",col="blue",lwd=2,type="n")
#points(myz,f_z,col="brown",pch=21,bg="brown")
#f_interp <- splinefun(myz, f_z, method = "fmm")
#lines(myz,f_interp(myz),col="magenta")

library(ggplot2)

# 1. Prepare data for plotting
# Ensure f_z is a numeric vector (in case it is a list from foreach)
if (is.list(f_z)) {
  # If it's a list of results, extract the 'mean' component
  if (is.list(f_z[[1]]) && "mean" %in% names(f_z[[1]])) {
    f_z_vec <- unlist(lapply(f_z, function(x) x$mean))
  } else {
    f_z_vec <- unlist(f_z)
  }
} else {
  f_z_vec <- f_z
}

# Create a smooth interpolation for the line
z_grid <- seq(min(myz), max(myz), length.out = 500)
f_interp <- splinefun(myz, f_z_vec, method = "fmm")

# Prepare data frames
plot_data <- data.frame(z = as.numeric(myz), f = as.numeric(f_z_vec))
line_data <- data.frame(z = as.numeric(z_grid), f = as.numeric(f_interp(z_grid)))

# 2. Create the plot
p <- ggplot(plot_data, aes(x = z, y = f)) +
  # Draw the interpolated smooth line from the line_data frame
  geom_line(data = line_data, color = "#8E44AD", linewidth = 1, alpha = 0.8) +
  # Draw the actual data points from the plot_data frame (inherited)
  geom_point(fill = "#D35400", color = "white", shape = 21, size = 3, stroke = 0.5) +
  # Labels and Title
  labs(
    title = "Marginal Posterior Density Estimation",
    subtitle = "Calculated via Vegas Integration with Spline Interpolation",
    x = "Parameter Space (z)",
    y = "Density",
    caption = "Source: Vegas R Integration"
  ) +
  # Apply a professional theme
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30", size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 20, 20, 20)
  )

```

```{r plot_marg2, include=TRUE, eval=TRUE, fig.width=10, fig.height=6, out.width="100%", fig.align='center'}
# 3. Display the plot
# plotting code above in hidden chunk
print(p)

```

