---
title: "Comparing Stan v Vegasr - Basket Trial"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Comparing Stan v Vegasr - Basket Trial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, eval=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# options(pkgdown.parallel_vignettes = FALSE)
# Check if running on GitHub Actions
# This is important as otherwise Github will attempt to build this and it's missing the python libraries and will fail
on_github <- Sys.getenv("GITHUB_ACTIONS") == "true"

   
# If on GitHub, set eval = FALSE for all subsequent chunks
knitr::opts_chunk$set(eval = !on_github)

knitr::opts_chunk$set(eval=TRUE)

```

```{r cran_check_setting,include=FALSE}
# this is to de-tune all examples to get past cran checks
#is_checking <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))
#if (is_checking) {
#     n_cores <- 2 # number of cores for computing marginal density on cran
#     no_den_pts<-5 # very ligth grid for on cran
#    } else { # running locally to go higher
#    n_cores <- parallel::detectCores() - 1
#no_den_pts=50
#}

```

## Quickstart
This article is a more complex version of the vegasr vignette [Rcpp](https://fraseriainlewis.github.io/vegasr/articles/rcpp.html) where the hierarchical Bayesian model now being estimated is a basket trial comprising of five baskets and binary response with logistic link function. This model has 14 parameters. This article includes a comparison with results from rstan. 


## Model Formulation

The model implemented is given below where $k$ denotes the basket, with $\beta_{0,k}$ and $\beta_{1,k}$ basket specific intercept and slope terms.

$$
\begin{aligned}
\mu_0 &\sim \text{Normal}(0, 2.5)\\
\sigma_0 &\sim \text{Half-Normal}(0, 2.5) \\
\mu_1 &\sim \text{Normal}(0, 2.5) \\
\sigma_1 &\sim \text{Half-Normal}(0, 2.5) \\
\beta_{0,k} &\sim \text{Normal}(\mu_0, \sigma_0)\hspace{0.95cm}\text{for}\enspace{k=0,\dots,4} \\
\beta_{1,k} &\sim \text{Normal}(\mu_1, \sigma_1) \hspace{0.95cm}\text{for}\enspace{k=0,\dots,4}\\
\text{logit(}{p_{i}}) &= \beta_{0,k} + \beta_{1,k} z_i\hspace{1.49cm}\text{for}\enspace{i=1,\dots,N}  \\
y_i &\sim \text{Bernoulli}(p_i)
\end{aligned}
$$


## Data set
The R chunk below creates a simple dataset of three cols:

-   a binary response variable y (0/1 = non-responder/responder),
-   a basket ID variable (1,2,3,4,5)
-   a binary treatment variable (0/1 = control/test treatment)

The basket ID is currently set fixed at 1, denoting there is only one basket in this trial, i.e. it's a classical two arm randomized trial design.

```{r datasets, include=TRUE}
library(vegasr)
library(loo)
library(parallel)
vegas_initialize()
thedata<-vegasr:::fn_create_data_5(99999) # a list of y and treat as matrices
# this function is in vegasr/R/fn_internal.R
```

## Fit Model via MCMC using RStan
Below are code chunks that implement the above model in Stan using a non-centralised parameterization (as recommended for Hamiltonian samplers).

```{r rsetup, include=FALSE}
library(rstan)
library(bayesplot)
library(ggplot2)
library(vegasr)
```

```{r stanmodeldef,include=TRUE,echo=TRUE}
### Stan model
  BHM_stan_1 <- "data {
  int<lower = 0> K;                     // number of baskets (K >= 2)
  int<lower = 0> N;                     // total number of participants
  //int<lower = 0, upper = N> N_k[K];     // K x 1 vector of basket sample sizes
  //int<lower = 0, upper = N> N_k;
  int<lower = 1, upper = K> k_vec[N];   // N x 1 vector of basket indicators
  int<lower = 0, upper = 1> z_vec[N];   // N x 1 vector of treatment indicators for active treatment arm
  int<lower = 0, upper = 1> y[N];       // N x 1 vector of binary responses
}
parameters {
  matrix[K,2] beta_tr;                  // K x 2 matrix of transformed regression coefficients
  real mu1;                              // scalar of hierarchical mean (log odds ratio)
  real<lower = 0> sigma1;                  // scalar of hierarchical SD (log odds ratio)
  real mu0;                              // scalar of hierarchical mean (log odds ratio)
  real<lower = 0> sigma0;                  // scalar of hierarchical SD (log odds ratio)

}
transformed parameters {
  matrix[K,2] beta;                     // K x 2 matrix of regression coefficients (without transformation)
  // Calculate beta coefs using transformed betas (leads to better mixing)
  for (k in 1:K){
    beta[k,1] = mu0 + sigma0 * beta_tr[k,1];
    beta[k,2] = mu1 + sigma1 * beta_tr[k,2];
  }

}
model {
  mu0 ~ normal(0, 2.5);    // vectorized normal priors for hierarchical means (specify SD in normal distn)
  sigma0 ~ normal(0, 2.5);       // vectorized half-normal priors for hierarchical SDs (specify SD in normal distn)
  mu1 ~ normal(0, 2.5);    // vectorized normal priors for hierarchical means (specify SD in normal distn)
  sigma1 ~ normal(0, 2.5);       // vectorized half-normal priors for hierarchical SDs (specify SD in normal distn)
  beta_tr[,1] ~ normal(0, 1);  // vectorized normal prior for the log odds for the control arm
  beta_tr[,2] ~ normal(0, 1);  // vectorized normal prior for the log odds ratios
  for (i in 1:N)
    y[i] ~ bernoulli_logit(beta[k_vec[i], 1] + beta[k_vec[i], 2] * z_vec[i]);
}
"

```

```{r stansample,echo=TRUE,results='hide'}

# Create list with input values for Stan model
data_input_1 <- list(
  K = length(unique(thedata$basket)),                # number of subgroups
  N = length(thedata$basket),                # total sample size
  #N_k = N-k,            # sample size per basket
  k_vec = as.integer(thedata$basket),        # N x 1 vector of subgroup indicators
  z_vec =as.integer(thedata$treat),        # N x 1 vector of active treatment arm indicators
  y = as.integer(thedata$y)                 # N x 1 vector of binary responses
  )


### Compile and fit model
options(mc.cores = parallel::detectCores())
# Compile model (only run the line below once for a simulation study - compilation not dependent on any
# simulation inputs as defined by scenarios or on simulated data)
stan_mod_1 <- stan_model(model_code = BHM_stan_1)
# Fit model
nsamps <- 150000       # number of posterior samples (after removing burn-in) per chain
nburnin <- 50000       # number of burn-in samples to remove at beginning of each chain
nchains <- 8          # number of chains
#BHM_pars_1 <- c("beta","mu0", "sigma0","mu1", "sigma1", "beta","beta_tr")     # parameters to sample
start_time <- Sys.time()
stan_fit_2 <- sampling(stan_mod_1, data = data_input_1,
                       iter = nsamps + nburnin, warmup = nburnin, chains = nchains)
end_time <- Sys.time()
end_time - start_time
post_draws_2 <- as.matrix(stan_fit_2)           # posterior samples of each parameter

```
```{r stanresults2,fig.width=10, fig.height=6, out.width="100%"}
library(bayesplot)
color_scheme_set("viridisA")
p<-mcmc_trace(stan_fit_2,"beta[1,1]")
#ggsave("precomputed/vg2_traceplot1_stan.png", p)
print(p)
print(summary(stan_fit_2))
```

## Example of calling log posterior functions
These functions are available in the source package in fns_arma.cpp or fns_eigen.cpp in /src.

```{r logpos,include=TRUE}
### Call the Rcpp functions to demonstrate correct input and output values

#print(mcmc_sum<-summary(stan_fit_2))
mcmc_sum<-summary(stan_fit_2)
smres<-mcmc_sum[[1]][11:24,c(4,8,9)]
print(round(smres,3))

```

## Log Evidence
As in the Rcpp vignette estimating the log evidence is not essential but we compute for completeness, also to a high precision. 

```{r logpos2,include=TRUE}

Rcpp::sourceCpp("/Users/work/vegasr/src/testing/fns_imp_eigen.cpp")

library(vegasr)
## Important - run next line in each R session to ensure python is ready
vegas_initialize()

K <- length(unique(thedata$basket))
lower <- c(rep(-0.9999, 2*K), -0.9999, -0.9999, 1e-4, 1e-4)
upper <- c(rep( 0.9999, 2*K),  0.9999,  0.9999,   0.9999,   0.9999)

result <- vegasBayesEvidence(
  f = vegasr::eigen_fn_log_post_2_par,
  lower = lower, upper = upper,
  nitn_warm = 1, neval_warm = 5000000,
  nitn = 1, neval = 5000000,
  errTol = 0.1, maxIter = 10, seed = 99999, nsearch = 100000,
  extra_args=list(
    y=thedata$y,
    treat=thedata$treat,
    basket = thedata$basket,
    shiftby=0,uselog=1.)
)

```

```{r logpos3,include=TRUE}
y_rv<-matrix(data=runif(5000000*length(result$x_grid)),ncol=length(result$x_grid))

# get X and Jac
res<-eigen_gridM(result$x_grid,
                 y_rv# this is just runif(), using runif from python currently
)
myX<-res[[1]]
myJac<-res[[2]]
# now compute f(x) and weights - note using shiftby and uselog, so need to revert those
f_x<-vegasr:::eigen_fn_log_post_2_par(myX,
                                      y=thedata$y,
                                      treat=thedata$treat,
                                      basket = thedata$basket,
                                      shiftby=result$shiftby,uselog=1.)

wgts<-exp(result$shiftby+f_x)*myJac # need to add back in shiftby and then exponentiation

## 
if(TRUE){
threshold<-quantile(wgts,0.9999)
epsilon <- sum(wgts[wgts>threshold])/sum(wgts)  
print(epsilon)  

# these these weights in the tails
extreme_idx <-which(wgts>threshold)
# compare before and after removal
# also check the high weights not in tails - for upper tail go back from top

}




if(FALSE){
  log_wgts<-log(wgts)
options(mc.cores = parallel::detectCores() - 1 )
psis_result<-psis(log_wgts,r_eff=1)
wgts2<-exp(psis_result$log_weights) # PSIS smoothed at top
}
if(TRUE){
#wgts<-f_x*myJac
#cliptop<-which(wgts>quantile(wgts,0.9999)) #
desc<-summary(wgts)
maxwgt<-desc["3rd Qu."]+100*(desc["3rd Qu."]-desc["1st Qu."])
#wgts2<-pmin(pmax(wgts, 0), quantile(wgts,0.9999)) # clip negative and top 0.01%
dropme<-which(wgts>maxwgt)
wgts2<-wgts[-dropme]
#wgts2<-wgts[-cliptop]
#myX<-myX[-cliptop,]
myX<-myX[-dropme,]
#plot(wgts2)
#wgts2<-wgts
wgts_std <- wgts2/sum(wgts2) # standardize
}
#wgts_std[422057]<-mean(wgts_std[-422057]);
#wgts_std<-pmin(pmax(wgts_std, 0), 1) # in case weight -ve, should not happen
#wgts_std <- wgts_std/sum(wgts_std) # standardize

cat("ESS = ",1/sum(wgts_std^2),"\n")

#### quantiles
X1<-myX[,1]
X1.srt<-sort(X1)
wgts_std.srt<-wgts_std[order(X1)]
idx1<-which(cumsum(wgts_std.srt)>0.025)[1]
idx2<-which(cumsum(wgts_std.srt)>0.975)[1]
p25raw<-X1.srt[idx1];
p25<-p25raw/(1-p25raw^2)
p975raw<-X1.srt[idx2];
p975<-p975raw/(1-p975raw^2)
cat("beta[1,1] 2.5%=",p25, " 97.5%=",p975,"\n")

X1<-myX[,1]
X1.srt<-sort(X1)
wgts_std.srt<-wgts_std[order(X1)]
idx1<-which(cumsum(wgts_std.srt)>0.025)[1]
idx2<-which(cumsum(wgts_std.srt)>0.975)[1]
p25raw<-X1.srt[idx1];
p25<-p25raw/(1-p25raw^2)
p975raw<-X1.srt[idx2];
p975<-p975raw/(1-p975raw^2)
cat("beta[1,1] 2.5%=",p25, " 97.5%=",p975,"\n")

X1<-myX[,2]
X1.srt<-sort(X1)
wgts_std.srt<-wgts_std[order(X1)]
idx1<-which(cumsum(wgts_std.srt)>0.025)[1]
idx2<-which(cumsum(wgts_std.srt)>0.975)[1]
p25raw<-X1.srt[idx1];
p25<-p25raw/(1-p25raw^2)
p975raw<-X1.srt[idx2];
p975<-p975raw/(1-p975raw^2)
cat("beta[2,1] 2.5%=",p25, " 97.5%=",p975,"\n")

X1<-myX[,3]
X1.srt<-sort(X1)
wgts_std.srt<-wgts_std[order(X1)]
idx1<-which(cumsum(wgts_std.srt)>0.025)[1]
idx2<-which(cumsum(wgts_std.srt)>0.975)[1]
p25raw<-X1.srt[idx1];
p25<-p25raw/(1-p25raw^2)
p975raw<-X1.srt[idx2];
p975<-p975raw/(1-p975raw^2)
cat("beta[3,1] 2.5%=",p25, " 97.5%=",p975,"\n")

X1<-myX[,11]
X1.srt<-sort(X1)
wgts_std.srt<-wgts_std[order(X1)]
idx1<-which(cumsum(wgts_std.srt)>0.025)[1]
idx2<-which(cumsum(wgts_std.srt)>0.975)[1]
p25raw<-X1.srt[idx1];
p25<-p25raw/(1-p25raw^2)
p975raw<-X1.srt[idx2];
p975<-p975raw/(1-p975raw^2)
cat("mu0 2.5%=",p25, " 97.5%=",p975,"\n")

X1<-myX[,13]
X1.srt<-sort(X1)
wgts_std.srt<-wgts_std[order(X1)]
idx1<-which(cumsum(wgts_std.srt)>0.025)[1]
idx2<-which(cumsum(wgts_std.srt)>0.975)[1]
p25raw<-X1.srt[idx1];
p25<-p25raw/(1-p25raw^2)
p975raw<-X1.srt[idx2];
p975<-p975raw/(1-p975raw^2)
cat("sd0 2.5%=",p25, " 97.5%=",p975,"\n")


```



