---
title: "Comparing Stan v Vegasr - Basket Trial"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Comparing Stan v Vegasr - Basket Trial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, eval=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# options(pkgdown.parallel_vignettes = FALSE)
# Check if running on GitHub Actions
# This is important as otherwise Github will attempt to build this and it's missing the python libraries and will fail
on_github <- Sys.getenv("GITHUB_ACTIONS") == "true"

   
# If on GitHub, set eval = FALSE for all subsequent chunks
knitr::opts_chunk$set(eval = !on_github)

knitr::opts_chunk$set(eval=TRUE)

```

```{r cran_check_setting,include=FALSE}
# this is to de-tune all examples to get past cran checks
#is_checking <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))
#if (is_checking) {
#     n_cores <- 2 # number of cores for computing marginal density on cran
#     no_den_pts<-5 # very ligth grid for on cran
#    } else { # running locally to go higher
#    n_cores <- parallel::detectCores() - 1
#no_den_pts=50
#}

```

## Quickstart
This article is a more complex version of the vegasr vignette [Rcpp](https://fraseriainlewis.github.io/vegasr/articles/rcpp.html) where the hierarchical Bayesian model now being estimated is a basket trial comprising of five baskets and binary response with logistic link function. This model has 14 parameters. This article includes a comparison with results from rstan. 


## Model Formulation

The model implemented is given below where $k$ denotes the basket, with $\beta_{0,k}$ and $\beta_{1,k}$ basket specific intercept and slope terms.

$$
\begin{aligned}
\mu_0 &\sim \text{Normal}(0, 2.5)\\
\sigma_0 &\sim \text{Half-Normal}(0, 2.5) \\
\mu_1 &\sim \text{Normal}(0, 2.5) \\
\sigma_1 &\sim \text{Half-Normal}(0, 2.5) \\
\beta_{0,k} &\sim \text{Normal}(\mu_0, \sigma_0)\hspace{0.95cm}\text{for}\enspace{k=0,\dots,4} \\
\beta_{1,k} &\sim \text{Normal}(\mu_1, \sigma_1) \hspace{0.95cm}\text{for}\enspace{k=0,\dots,4}\\
\text{logit(}{p_{i}}) &= \beta_{0,k} + \beta_{1,k} z_i\hspace{1.49cm}\text{for}\enspace{i=1,\dots,N}  \\
y_i &\sim \text{Bernoulli}(p_i)
\end{aligned}
$$


## Data set
The R chunk below creates a simple dataset of three cols:

-   a binary response variable y (0/1 = non-responder/responder),
-   a basket ID variable (1,2,3,4,5)
-   a binary treatment variable (0/1 = control/test treatment)

The basket ID is currently set fixed at 1, denoting there is only one basket in this trial, i.e. it's a classical two arm randomized trial design.

```{r datasets, include=TRUE}
library(vegasr)
library(loo)
library(parallel)
vegas_initialize()
thedata<-vegasr:::fn_create_data_5(99999) # a list of y and treat as matrices
# this function is in vegasr/R/fn_internal.R
```

## Fit Model via MCMC using RStan
Below are code chunks that implement the above model in Stan using a non-centralised parameterization (as recommended for Hamiltonian samplers).

```{r rsetup, include=FALSE}
library(rstan)
library(bayesplot)
library(ggplot2)
library(vegasr)
```

```{r stanmodeldef,include=TRUE,echo=TRUE}
### Stan model
  BHM_stan_1 <- "data {
  int<lower = 0> K;                     // number of baskets (K >= 2)
  int<lower = 0> N;                     // total number of participants
  //int<lower = 0, upper = N> N_k[K];     // K x 1 vector of basket sample sizes
  //int<lower = 0, upper = N> N_k;
  int<lower = 1, upper = K> k_vec[N];   // N x 1 vector of basket indicators
  int<lower = 0, upper = 1> z_vec[N];   // N x 1 vector of treatment indicators for active treatment arm
  int<lower = 0, upper = 1> y[N];       // N x 1 vector of binary responses
}
parameters {
  matrix[K,2] beta_tr;                  // K x 2 matrix of transformed regression coefficients
  real mu1;                              // scalar of hierarchical mean (log odds ratio)
  real<lower = 0> sigma1;                  // scalar of hierarchical SD (log odds ratio)
  real mu0;                              // scalar of hierarchical mean (log odds ratio)
  real<lower = 0> sigma0;                  // scalar of hierarchical SD (log odds ratio)

}
transformed parameters {
  matrix[K,2] beta;                     // K x 2 matrix of regression coefficients (without transformation)
  // Calculate beta coefs using transformed betas (leads to better mixing)
  for (k in 1:K){
    beta[k,1] = mu0 + sigma0 * beta_tr[k,1];
    beta[k,2] = mu1 + sigma1 * beta_tr[k,2];
  }

}
model {
  mu0 ~ normal(0, 2.5);    // vectorized normal priors for hierarchical means (specify SD in normal distn)
  sigma0 ~ normal(0, 2.5);       // vectorized half-normal priors for hierarchical SDs (specify SD in normal distn)
  mu1 ~ normal(0, 2.5);    // vectorized normal priors for hierarchical means (specify SD in normal distn)
  sigma1 ~ normal(0, 2.5);       // vectorized half-normal priors for hierarchical SDs (specify SD in normal distn)
  beta_tr[,1] ~ normal(0, 1);  // vectorized normal prior for the log odds for the control arm
  beta_tr[,2] ~ normal(0, 1);  // vectorized normal prior for the log odds ratios
  for (i in 1:N)
    y[i] ~ bernoulli_logit(beta[k_vec[i], 1] + beta[k_vec[i], 2] * z_vec[i]);
}
"

```

```{r stansample,echo=TRUE,results='hide'}

# Create list with input values for Stan model
data_input_1 <- list(
  K = length(unique(thedata$basket)),                # number of subgroups
  N = length(thedata$basket),                # total sample size
  #N_k = N-k,            # sample size per basket
  k_vec = as.integer(thedata$basket),        # N x 1 vector of subgroup indicators
  z_vec =as.integer(thedata$treat),        # N x 1 vector of active treatment arm indicators
  y = as.integer(thedata$y)                 # N x 1 vector of binary responses
  )


### Compile and fit model
options(mc.cores = parallel::detectCores())
# Compile model (only run the line below once for a simulation study - compilation not dependent on any
# simulation inputs as defined by scenarios or on simulated data)
stan_mod_1 <- stan_model(model_code = BHM_stan_1)
# Fit model
nsamps <- 10000       # number of posterior samples (after removing burn-in) per chain
nburnin <- 5000       # number of burn-in samples to remove at beginning of each chain
nchains <- 4          # number of chains
#BHM_pars_1 <- c("beta","mu0", "sigma0","mu1", "sigma1", "beta","beta_tr")     # parameters to sample
start_time <- Sys.time()
stan_fit_2 <- sampling(stan_mod_1, data = data_input_1,
                       iter = nsamps + nburnin, warmup = nburnin, chains = nchains)
end_time <- Sys.time()
end_time - start_time
post_draws_2 <- as.matrix(stan_fit_2)           # posterior samples of each parameter

```
```{r stanresults2,fig.width=10, fig.height=6, out.width="100%"}
library(bayesplot)
color_scheme_set("viridisA")
p<-mcmc_trace(stan_fit_2,"beta[1,1]")
#ggsave("precomputed/vg2_traceplot1_stan.png", p)
print(p)
print(summary(stan_fit_2))
```

## Example of calling log posterior functions
These functions are available in the source package in fns_arma.cpp or fns_eigen.cpp in /src.

```{r logpos,include=TRUE}
### Call the Rcpp functions to demonstrate correct input and output values

#print(mcmc_sum<-summary(stan_fit_2))
print(stan_fit_2)
#smres<-mcmc_sum[[1]][11:24,c(4,8,9)]
#print(round(smres,3))

stansum<-summary(stan_fit_2)$summary

my25Stan<-stansum[c("mu1","sigma1","mu0","sigma0","beta[1,1]","beta[1,2]","beta[2,1]","beta[2,2]","beta[3,1]","beta[3,2]","beta[4,1]",   "beta[4,2]","beta[5,1]","beta[5,2]"),4]

```

## Compute quantiles for all variables
Do this in one step as no more computation needed. gotcha is dealing with extreme weights. tbd.

```{r logpos2,include=TRUE}
#############################################################################################
# 1. Run vegas to compute integrand - log evidence and then manually use importance sampling
# from the same grid and compare result. uses different random samples but same grid (weights)
# so should be same up to monte-carlo error.

K <- length(unique(thedata$basket))
lower <- c(rep(-0.9999, 2*K), -0.9999, -0.9999, 1e-4, 1e-4)
upper <- c(rep( 0.9999, 2*K),  0.9999,  0.9999,   0.9999,   0.9999)

result <- vegasBayesEvidence(
  f = vegasr::eigen_fn_log_post_2_par,
  lower = lower, upper = upper,
  nitn_warm = 10, neval_warm = 5000000,
  nitn = 10, neval = 1000000,
  errTol = 0.1, maxIter = 20, seed = 999199, nsearch = 100000,
  extra_args=list(
    y=thedata$y,
    treat=thedata$treat,
    basket = thedata$basket,
    shiftby=0,uselog=1.)
)

cat("log evidence direct from vegas = ",result$log_evidence,"\n")

#
f_x<-vegasr:::eigen_fn_log_post_2_par(result$x_vals, # these are random samples from the final grid
                                       y=thedata$y,
                                       treat=thedata$treat,
                                       basket = thedata$basket,
                                       shiftby=result$shiftby, # need to use same offset as above
                                       uselog=1. # use log as just function eval
                                      )

# result$pwgts - these are sampling weights - the internal jacobian and volume adjustment
# this is usual importance sampling for integrand - function value times weight
# this weight has n in it so it's sum below not mean (but conceptually it's mean)
log_ev2<-log(sum(exp(result$shiftby+f_x)*result$pwgts)) # adjust by offset,

cat("log evidence resampled from vegas = ",log_ev2,"\n")


#############################################################################################
# 2. Use the final grid to estimate marginal quantiles

# generate
y_rv<-matrix(data=runif(10000000*length(result$x_grid)),ncol=length(result$x_grid))

# get X and Jac
# note - this needs moved to multi-threaded
res<-vegasr:::eigen_gridM(result$x_grid,
                 y_rv# this is just runif(), using runif from python currently
)
myX<-res[[1]]
myJac<-res[[2]]
# now compute f(x) and weights - note using shiftby and uselog, so need to revert those
f_x<-vegasr:::eigen_fn_log_post_2_par(myX,
                                      y=thedata$y,
                                      treat=thedata$treat,
                                      basket = thedata$basket,
                                      shiftby=result$shiftby,uselog=1.)


wgts<-exp(result$shiftby+f_x)*myJac # need to add back in shiftby and then exponentiation
idx<-which(wgts>0)
wgts<-wgts[idx]
myX<-myX[idx,]

#wgts<-pmin(pmax(wgts, 0), quantile(wgts,0.9999))
wgts_std <- wgts/sum(wgts) # standardize
#length(which(wgts_std==0))/length(wgts)

quantile(sample(myX[,14],size=1E6,prob = wgts_std,replace=TRUE),0.025)


## from here we have all the info we need to compute all the quantiles for all variables with no more computation
# consider X1
myorder1<-order(myX[,1]) # sort asc idx
X1o<-myX[myorder1,1] # sort X1 asc
w1<-wgts_std[myorder1] # order weights
w1b<-cumsum(w1); # cumsum weights
p25X<-X1o[(which(w1b>0.025)[1])] # get 2.5% on x scale -1,+1
p25Z<-p25X/(1-p25X^2) # transform back to -inf, + inf

cat("Variable beta[1,1] = 2.5% ",p25Z,"\n")

getperc<-function(X.loc,idx.loc,p.loc,wgts.loc){
  myorder<-order(X.loc[,idx.loc]) # sort asc idx
  Xo<-myX[myorder,idx.loc] # sort X1 asc
  w1<-wgts.loc[myorder] # order weights
  w1b<-cumsum(w1); # cumsum weights
  pX<-Xo[(which(w1b>p.loc)[1])] # get 2.5% on x scale -1,+!
  pZ<-pX/(1-pX^2) # transform back to -inf, + inf
  return(pZ)
}

my25<-rep(0,14)
names(my25)<-c("beta[1,1]","beta[2,1]","beta[3,1]","beta[4,1]","beta[5,1]",
               "beta[1,2]","beta[2,2]","beta[3,2]","beta[4,2]","beta[5,2]",
               "mu0","mu1","sigma0","sigma1")
for(i in 1:14){my25[i]<-getperc(myX,i,0.025,wgts_std)}


my25<-my25[c("mu1","sigma1","mu0","sigma0","beta[1,1]","beta[1,2]","beta[2,1]","beta[2,2]","beta[3,1]","beta[3,2]","beta[4,1]",   "beta[4,2]","beta[5,1]","beta[5,2]")]


m<-cbind(my25Stan,my25,100*(my25Stan-my25)/my25Stan)
colnames(m)<-c("stan","vegas","% diff")
print(m)
```



