---
title: "Comparing Stan v Vegasr - Basket Trial"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Comparing Stan v Vegasr - Basket Trial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, eval=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# options(pkgdown.parallel_vignettes = FALSE)
# Check if running on GitHub Actions
# This is important as otherwise Github will attempt to build this and it's missing the python libraries and will fail
on_github <- Sys.getenv("GITHUB_ACTIONS") == "true"

   
# If on GitHub, set eval = FALSE for all subsequent chunks
knitr::opts_chunk$set(eval = !on_github)

knitr::opts_chunk$set(eval=TRUE)

```

```{r cran_check_setting,include=FALSE}
# this is to de-tune all examples to get past cran checks
#is_checking <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))
#if (is_checking) {
#     n_cores <- 2 # number of cores for computing marginal density on cran
#     no_den_pts<-5 # very ligth grid for on cran
#    } else { # running locally to go higher
#    n_cores <- parallel::detectCores() - 1
#no_den_pts=50
#}

```

## Quickstart
This article is a more complex version of the vegasr vignette [Rcpp](https://fraseriainlewis.github.io/vegasr/articles/rcpp.html) where the hierarchical Bayesian model now being estimated is a basket trial comprising of five baskets and binary response with logistic link function. This model has 14 parameters. This article includes a comparison with results from rstan. 


## Model Formulation

The model implemented is given below where $k$ denotes the basket, with $\beta_{0,k}$ and $\beta_{1,k}$ basket specific intercept and slope terms.

$$
\begin{aligned}
\mu_0 &\sim \text{Normal}(0, 2.5)\\
\sigma_0 &\sim \text{Half-Normal}(0, 2.5) \\
\mu_1 &\sim \text{Normal}(0, 2.5) \\
\sigma_1 &\sim \text{Half-Normal}(0, 2.5) \\
\beta_{0,k} &\sim \text{Normal}(\mu_0, \sigma_0)\hspace{0.95cm}\text{for}\enspace{k=0,\dots,4} \\
\beta_{1,k} &\sim \text{Normal}(\mu_1, \sigma_1) \hspace{0.95cm}\text{for}\enspace{k=0,\dots,4}\\
\text{logit(}{p_{i}}) &= \beta_{0,k} + \beta_{1,k} z_i\hspace{1.49cm}\text{for}\enspace{i=1,\dots,N}  \\
y_i &\sim \text{Bernoulli}(p_i)
\end{aligned}
$$


## Data set
The R chunk below creates a simple dataset of three cols:

-   a binary response variable y (0/1 = non-responder/responder),
-   a basket ID variable (1,2,3,4,5)
-   a binary treatment variable (0/1 = control/test treatment)

The basket ID is currently set fixed at 1, denoting there is only one basket in this trial, i.e. it's a classical two arm randomized trial design.

```{r datasets, include=TRUE}
library(vegasr)
vegas_initialize()
thedata<-vegasr:::fn_create_data_5(99999) # a list of y and treat as matrices
# this function is in vegasr/R/fn_internal.R
```

## Fit Model via MCMC using RStan
Below are code chunks that implement the above model in Stan using a non-centralised parameterization (as recommended for Hamiltonian samplers).

```{r rsetup, include=FALSE}
library(rstan)
library(bayesplot)
library(ggplot2)
library(vegasr)
```

```{r stanmodeldef,include=TRUE,echo=TRUE}
### Stan model
  BHM_stan_1 <- "data {
  int<lower = 0> K;                     // number of baskets (K >= 2)
  int<lower = 0> N;                     // total number of participants
  //int<lower = 0, upper = N> N_k[K];     // K x 1 vector of basket sample sizes
  //int<lower = 0, upper = N> N_k;
  int<lower = 1, upper = K> k_vec[N];   // N x 1 vector of basket indicators
  int<lower = 0, upper = 1> z_vec[N];   // N x 1 vector of treatment indicators for active treatment arm
  int<lower = 0, upper = 1> y[N];       // N x 1 vector of binary responses
}
parameters {
  matrix[K,2] beta_tr;                  // K x 2 matrix of transformed regression coefficients
  real mu1;                              // scalar of hierarchical mean (log odds ratio)
  real<lower = 0> sigma1;                  // scalar of hierarchical SD (log odds ratio)
  real mu0;                              // scalar of hierarchical mean (log odds ratio)
  real<lower = 0> sigma0;                  // scalar of hierarchical SD (log odds ratio)

}
transformed parameters {
  matrix[K,2] beta;                     // K x 2 matrix of regression coefficients (without transformation)
  // Calculate beta coefs using transformed betas (leads to better mixing)
  for (k in 1:K){
    beta[k,1] = mu0 + sigma0 * beta_tr[k,1];
    beta[k,2] = mu1 + sigma1 * beta_tr[k,2];
  }

}
model {
  mu0 ~ normal(0, 2.5);    // vectorized normal priors for hierarchical means (specify SD in normal distn)
  sigma0 ~ normal(0, 2.5);       // vectorized half-normal priors for hierarchical SDs (specify SD in normal distn)
  mu1 ~ normal(0, 2.5);    // vectorized normal priors for hierarchical means (specify SD in normal distn)
  sigma1 ~ normal(0, 2.5);       // vectorized half-normal priors for hierarchical SDs (specify SD in normal distn)
  beta_tr[,1] ~ normal(0, 1);  // vectorized normal prior for the log odds for the control arm
  beta_tr[,2] ~ normal(0, 1);  // vectorized normal prior for the log odds ratios
  for (i in 1:N)
    y[i] ~ bernoulli_logit(beta[k_vec[i], 1] + beta[k_vec[i], 2] * z_vec[i]);
}
"

```

```{r stansample,echo=TRUE,results='hide'}

# Create list with input values for Stan model
data_input_1 <- list(
  K = length(unique(thedata$basket)),                # number of subgroups
  N = length(thedata$basket),                # total sample size
  #N_k = N-k,            # sample size per basket
  k_vec = as.integer(thedata$basket),        # N x 1 vector of subgroup indicators
  z_vec =as.integer(thedata$treat),        # N x 1 vector of active treatment arm indicators
  y = as.integer(thedata$y)                 # N x 1 vector of binary responses
  )


### Compile and fit model
options(mc.cores = parallel::detectCores())
# Compile model (only run the line below once for a simulation study - compilation not dependent on any
# simulation inputs as defined by scenarios or on simulated data)
stan_mod_1 <- stan_model(model_code = BHM_stan_1)
# Fit model
nsamps <- 50000       # number of posterior samples (after removing burn-in) per chain
nburnin <- 5000       # number of burn-in samples to remove at beginning of each chain
nchains <- 4          # number of chains
#BHM_pars_1 <- c("beta","mu0", "sigma0","mu1", "sigma1", "beta","beta_tr")     # parameters to sample
start_time <- Sys.time()
stan_fit_2 <- sampling(stan_mod_1, data = data_input_1,
                       iter = nsamps + nburnin, warmup = nburnin, chains = nchains)
end_time <- Sys.time()
end_time - start_time
post_draws_2 <- as.matrix(stan_fit_2)           # posterior samples of each parameter

```
```{r stanresults2,fig.width=10, fig.height=6, out.width="100%"}
library(bayesplot)
color_scheme_set("viridisA")
p<-mcmc_trace(stan_fit_2,"beta[1,1]")
#ggsave("precomputed/vg2_traceplot1_stan.png", p)
print(p)
print(summary(stan_fit_2))
```

## Example of calling log posterior functions
These functions are available in the source package in fns_arma.cpp or fns_eigen.cpp in /src.

```{r logpos,include=TRUE,eval=FALSE}
### Call the Rcpp functions to demonstrate correct input and output values

theta   <- matrix(data=rep(0.1,length=4*14), ncol = 14) # 5 intercept + 5 slope + 4 hyper
theta<-jitter(theta)

vegasr::eigen_fn_log_post_2(theta, thedata$y, thedata$treat, thedata$basket,0.0, 1.0)

```

## Log Evidence
As in the Rcpp vignette estimating the log evidence is not essential but we compute for completeness, also to a high precision. 

```{r logpos2,include=TRUE,eval=TRUE}
library(vegasr)
vegas_initialize()
library(tictoc)
K <- length(unique(thedata$basket))
lower <- c(rep(-0.9999, 2*K), -0.9999, -0.9999, 1e-4, 1e-4)
upper <- c(rep( 0.9999, 2*K),  0.9999,  0.9999,   0.9999,   0.9999)

tic()
result_logEv <- vegasBayesEvidence(
  f = vegasr::eigen_fn_log_post_2_par,
  lower = lower, upper = upper,
  nitn_warm = 10, neval_warm = 100000,
  nitn = 10, neval = 100000,
  errTol = 0.1, maxIter = 20, seed = 99999, nsearch = 100000,
  extra_args=list(
    y=thedata$y,
    treat=thedata$treat,
    basket = thedata$basket,
    shiftby=0,uselog=1.)
)
cat("log evidence = ",result_logEv,"\n")
toc()


```
## 2. Find Marginal
Example using parallel version of log posterior using Eigen library. See fns_eigen.cpp in /src in source package.

```{r marg_single,include=TRUE,eval=FALSE}
tic()
mymarg<-vegasBayesPosterior(f=vegasr::eigen_fn_marg_2,
                            lower=lower[-1],
                            upper=upper[-1],
                            nitn_warm = 10, neval_warm = 50000,
                            nitn = 10, neval = 50000,
                            errTol=1,maxIter=5,seed=99999,nsearch=10000,
                            log_evidence = result_logEv,
                            extra_args=list(
                              y=thedata$y,
                              treat=thedata$treat,
                              basket = thedata$basket,
                              shiftby=0,uselog=1.,z=-0.98))
cat("Marginal density f(z) at z = -0.1 = ",mymarg,"\n")
toc()

```

```{r marg_beta11,include=TRUE}
#result_logEv<-0.0
#myz<-c(seq(-1.7,-1.21,len=25),seq(-1.2,-0.7,len=50),seq(-0.69,-0.4,len=25))

library(vegasr)
vegas_initialize()
library(tictoc)
K <- length(unique(thedata$basket))
lower <- c(rep(-0.9999, 2*K), -0.9999, -0.9999, 1e-4, 1e-4)
upper <- c(rep( 0.9999, 2*K),  0.9999,  0.9999,   0.9999,   0.9999)


myz<-c(seq(-1.7,-0.4,len=20))

tic("") # Start timer with a label
f_z<-rep(0,length(myz));
i<-1;
for(z in myz){
  f_z[i]<-vegasBayesPosterior(f=vegasr::eigen_fn_marg_2,
                            lower=lower[-1],
                            upper=upper[-1],
                            nitn_warm = 20, neval_warm = 100000,
                            nitn = 10, neval = 200000,
                            errTol=1,maxIter=20,seed=99999,nsearch=10000,
                            log_evidence = result_logEv,
                            extra_args=list(
                              y=thedata$y,
                              treat=thedata$treat,
                              basket = thedata$basket,
                              shiftby=0,uselog=1.,z=z))
  cat("i=",i," z=",z," fz=",f_z[i],"\n")
  i<-i+1
}

toc() # Stops timer and prints

```

```{r marg_beta111,include=TRUE}
logun<-log(f_z)+result_logEv # back to unstand log density
f_z_un<-exp(logun) # back to unstand density
ff_interp <- splinefun(myz, as.matrix(f_z_un), method = "fmm") # fit fun to every point
evA<-integrate(ff_interp,min(myz),max(myz))$value # find area 
ff_z<-f_z_un/evA # divide by area to standardize marginal density
ff_interp <- splinefun(myz, as.matrix(ff_z), method = "fmm") # now

#plot(myz,predict(loess(ff_z ~ myz, span = 0.1,family="g",from=min(myz),to=max(myz))))

```

## Comparison with RStan
We compare one of the marginal densities, for $\beta_{0,0}$. These are very similar, as expected. One of the open questions is how to compare these two approaches in terms of numerical speed and accuracy, as both are approximations, and each gets better with more samples (e.g. n_eff in rstan output and neval parameter in vegas). 

```{r plot_marg, include=FALSE, fig.width=10, fig.height=6, out.width="100%", fig.align='center'}

#if(!nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))){
library(ggplot2)

# 1. Prepare data for plotting
# Ensure f_z is a numeric vector (in case it is a list from foreach)
#if (is.list(f_z)) {
#  # If it's a list of results, extract the 'mean' component
#  if (is.list(f_z[[1]]) && "mean" %in% names(f_z[[1]])) {
#    f_z_vec <- unlist(lapply(f_z, function(x) x$mean))
#  } else {
#    f_z_vec <- unlist(f_z)
#  }
#} else {
  f_z_vec <- f_z
  ff_z_vec <- ff_z
#}

# Create a smooth interpolation for the line
z_grid <- seq(min(myz), max(myz), length.out = 500)
f_interp <- splinefun(myz, f_z_vec, method = "fmm")

# Prepare data frames
#plot_data <- data.frame(z = as.numeric(myz), f = as.numeric(f_z_vec))
#line_data <- data.frame(z = as.numeric(z_grid), f = as.numeric(f_interp(z_grid)))

plot2_data <- data.frame(z = as.numeric(myz), f = as.numeric(ff_z_vec))
line2_data <- data.frame(z = as.numeric(z_grid), f = as.numeric(ff_interp(z_grid)))

stan_samples <- as.numeric(extract(stan_fit_2, "beta[1,1]")[["beta[1,1]"]])
stan_df <- data.frame(z = stan_samples)

p <- ggplot() +
  # --- STAN POSTERIOR DENSITY --- 
  geom_density(data = stan_df, aes(x = z, color = "Stan Posterior", fill = "Stan Posterior"), 
                  alpha = 0.2, linewidth = 1) +
  # --- FIRST DATASET (f_z) ---
  # Draw the interpolated smooth line
  #geom_line(data = line_data, aes(x = z, y = f, color = "Original Log Ev"), linewidth = 1, alpha = 0.8) +
  # Draw the actual data points
  #geom_point(data = plot_data, aes(x = z, y = f, fill = "Original Log Ev"), color = "white", shape = 21, size = 3, stroke = 0.5) +
  
  # --- SECOND DATASET (ff_z) ---
  # Draw the interpolated smooth line
  geom_line(data = line2_data, aes(x = z, y = f, color = "Marginal standardized"), linewidth = 1, alpha = 0.8) +
  # Draw the actual data points
  geom_point(data = plot2_data, aes(x = z, y = f, fill = "Marginal standardized"), color = "white", shape = 21, size = 3, stroke = 0.5) +
  
  # Custom Colors and Fills
  scale_color_manual(name = "Dataset", values = c( 
                                                  "Marginal standardized" = "red",
                                                  "Stan Posterior" = "blue")) +
  scale_fill_manual(name = "Dataset", values = c(
                                                 "Marginal standardized" = "orange",
                                                 "Stan Posterior" = "grey")) +
  
  # Labels and Title
  labs(
    title = "Marginal Posterior Density Estimation",
    subtitle = "Comparison of posterior densities",
    x = "Parameter Space (z)",
    y = "Density",
    caption = "Source: Vegas R Integration"
  ) +
  
  # Apply a professional theme
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30", size = 11),
    axis.title = element_text(face = "bold"),
    legend.position = "top",
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 20, 20, 20)
  )
#}
```

```{r plot_marg2, include=TRUE,  fig.width=10, fig.height=6, out.width="100%", fig.align='center'}
# 3. Display the plot
# plotting code above in hidden chunk
#if(!nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))){
print(p)
#}
```


